---
title: "Student Performance Factors Dashboard"
format: 
  dashboard:
    orientation: rows
    theme: flatly
    embed-resources: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(plotly)
library(scales)
library(DT)
library(ggplot2)

#Load data
df <- read.csv("StudentPerformanceFactors.csv")
```

## Row

### Project Overview
This dashboard explores the Student Performance Factors dataset from Kaggle, which contains information about students' academic performance and the personal and social-economical factors that may influence their Exam_Score.
The dataset includes variables such as:
Academic behavior: Hours_Studied, Attendance, Previous_Scores, Tutoring_Sessions
Personal/lifestyle factors: Sleep_Hours, Physical_Activity, Motivation_Level
Family and home context: Family_Income, Parental_Involvement, Parental_Education_Level, Internet_Access
School context: Teacher_Quality, School_Type, Access_to_Resources
Outcome variable: Exam_Score

### Project Goals
The goal of this project is to understand which factors appear most associated with student academic performance and to identify patterns that may help explain differences in exam results. Student success is influenced by more than intelligence alone. By exploring academic, social, and environmental factors together, this dashboard helps highlight which conditions may support stronger student outcomes.

## Row
## Static insight
```{r}
ggplot(df, aes(x = Exam_Score)) +
  geom_histogram(bins = 25, fill = "#2C7FB8", color = "white") +
  labs(
    title = "Distribution of Exam Scores",
    x = "Exam Score",
    y = "Number of Students"
  ) +
  theme_minimal(base_size = 13)
```


## Row 
### Interactive Visualization 1

```{r}
#| title: "U.S. Unemployment Over Time"
df$Family_Income        <- as.character(df$Family_Income)
df$Parental_Involvement <- as.character(df$Parental_Involvement)
df$Access_to_Resources  <- as.character(df$Access_to_Resources)

pi_levels  <- c("High", "Medium", "Low")
atr_levels <- c("High", "Medium", "Low")
pi_opts    <- c("All", pi_levels)
atr_opts   <- c("All", atr_levels)

income_colors <- c(
  "Low"    = "#EF553B",
  "Medium" = "#00CC96",
  "High"   = "#636EFA"
)

combos <- expand.grid(pi = pi_opts, atr = atr_opts, stringsAsFactors = FALSE)

fig            <- plot_ly()
visibility_map <- list()
trace_index    <- 0

for (i in seq_len(nrow(combos))) {
  pi_val     <- combos$pi[i]
  atr_val    <- combos$atr[i]
  is_default <- (pi_val == "All" && atr_val == "All")

  sub <- df
  if (pi_val  != "All") sub <- sub[sub$Parental_Involvement == pi_val, ]
  if (atr_val != "All") sub <- sub[sub$Access_to_Resources  == atr_val, ]

  sub$Family_Income <- factor(sub$Family_Income, levels = c("Low", "Medium", "High"))

  for (inc in c("Low", "Medium", "High")) {
    sub_inc <- sub[sub$Family_Income == inc, ]
    scores  <- sub_inc$Exam_Score

    # Pre-compute box stats
    mn  <- round(mean(scores, na.rm = TRUE), 2)
    q1  <- quantile(scores, 0.25, na.rm = TRUE)
    med <- median(scores, na.rm = TRUE)
    q3  <- quantile(scores, 0.75, na.rm = TRUE)
    iqr <- q3 - q1
    lo  <- round(min(scores, na.rm = TRUE), 2)
    hi  <- round(max(scores, na.rm = TRUE), 2)

    hover_text <- paste0(
      "<b>", inc, " Income</b><br>",
      "Max: ", hi,  "<br>",
      "Mean: ", mn, "<br>",
      "Median: ", med, "<br>",
      "Min: ", lo
    )

    fig <- fig %>% add_trace(
      x              = sub_inc$Family_Income,
      y              = scores,
      type           = "box",
      name           = inc,
      marker         = list(color = income_colors[[inc]]),
      line           = list(color = income_colors[[inc]]),
      fillcolor      = paste0(income_colors[[inc]], "55"),
      boxpoints      = "outliers",
      hoverinfo      = "none",          # disable box hover entirely
      visible        = is_default,
      legendgroup    = inc,
      showlegend     = (i == 0),
      offsetgroup    = inc,
      alignmentgroup = "fixed"
    )

    trace_index <- trace_index + 1
    visibility_map[[length(visibility_map) + 1]] <- list(
      pi = pi_val, atr = atr_val, trace = trace_index
    )

    # Invisible scatter point at median — carries the unified tooltip
    fig <- fig %>% add_trace(
      x          = inc,
      y          = med,
      type       = "scatter",
      mode       = "markers",
      name       = inc,
      legendgroup = inc,
      showlegend = FALSE,
      marker     = list(color = "rgba(0,0,0,0)", size = 40),  # fully transparent, large hit area
      hovertemplate = paste0(
        "<b>", inc, " Income</b><br>",
        "Max: ", hi,  "<br>",
        "Mean: ", mn, "<br>",
        "Median: ", med, "<br>",
        "Min: ", lo,
        "<extra></extra>"
      ),
      visible    = is_default
    )

    trace_index <- trace_index + 1
    visibility_map[[length(visibility_map) + 1]] <- list(
      pi = pi_val, atr = atr_val, trace = trace_index
    )
    
    # Outlier points with tooltips
    outliers <- scores[scores < q1 - 1.5 * iqr | scores > q3 + 1.5 * iqr]
    if (length(outliers) > 0) {
      fig <- fig %>% add_trace(
        x             = rep(inc, length(outliers)),
        y             = outliers,
        type          = "scatter",
        mode          = "markers",
        name          = inc,
        legendgroup   = inc,
        showlegend    = FALSE,
        marker        = list(color = income_colors[[inc]], size = 6),
        hoverlabel    = list(bgcolor = "#000000", font = list(color = "#ffffff")),
        hovertemplate = paste0(
          "<b>", inc, " Income</b><br>",
          "Outlier: %{y}<br>",
          "<extra></extra>"
        ),
        visible = is_default
      )

      trace_index <- trace_index + 1
      visibility_map[[length(visibility_map) + 1]] <- list(
        pi = pi_val, atr = atr_val, trace = trace_index
      )
    }
  }
}

make_vis <- function(sel_pi, sel_atr) {
  vapply(visibility_map, function(m) m$pi == sel_pi && m$atr == sel_atr, logical(1))
}

pi_buttons <- lapply(pi_opts, function(pv) {
  list(method = "update",
       args   = list(list(visible = as.list(make_vis(pv, "All")))),
       label  = pv)
})

atr_buttons <- lapply(atr_opts, function(av) {
  list(method = "update",
       args   = list(list(visible = as.list(make_vis("All", av)))),
       label  = av)
})

fig <- fig %>%
  layout(
    title = list(
      text    = "Effect of Family Income on Exam Performance",
      x       = 0.5,
      xanchor = "center",
      y       = 0.97,
      yanchor = "top",
      font    = list(size = 16)
    ),
    xaxis = list(
      title         = "Family Income",
      type          = "category",
      categoryorder = "array",
      categoryarray = c("Low", "Medium", "High")
    ),
    yaxis   = list(title = "Exam Score"),
    boxmode = "overlay",
    margin  = list(t = 130, r = 30, b = 60, l = 60),

    updatemenus = list(
      list(
        type        = "dropdown",
        active      = 0,
        x           = 0.01,
        xanchor     = "left",
        y           = 1.10,
        yanchor     = "top",
        buttons     = pi_buttons,
        bgcolor     = "#ffffff",
        bordercolor = "#aaaaaa",
        font        = list(size = 12)
      ),
      list(
        type        = "dropdown",
        active      = 0,
        x           = 0.13,
        xanchor     = "left",
        y           = 1.10,
        yanchor     = "top",
        buttons     = atr_buttons,
        bgcolor     = "#ffffff",
        bordercolor = "#aaaaaa",
        font        = list(size = 12)
      )
    ),

    annotations = list(
      list(
        text      = "<b>Parental Involvement:</b>",
        x = 0,  xref = "paper", xanchor = "left",
        y = 1.17, yref = "paper", yanchor = "top",
        showarrow = FALSE, font = list(size = 12, color = "#333333")
      ),
      list(
        text      = "<b>Access to Resources:</b>",
        x = 0.12, xref = "paper", xanchor = "left",
        y = 1.17, yref = "paper", yanchor = "top",
        showarrow = FALSE, font = list(size = 12, color = "#333333")
      )
    )
  )

fig
```

### Interactive visualization 2

```{r}
#| title: "U.S. Unemployment Over the last decade in the data"
#rename columns
df <- df %>%
  rename(
     HoursStudied = Hours_Studied,
     ExamScore = Exam_Score,
     Attendance = Attendance
   )

#filter outliers
q1_h <- quantile(df$HoursStudied, 0.25, na.rm = TRUE)
q3_h <- quantile(df$HoursStudied, 0.75, na.rm = TRUE)
iqr_h <- q3_h - q1_h

q1_s <- quantile(df$ExamScore, 0.25, na.rm = TRUE)
q3_s <- quantile(df$ExamScore, 0.75, na.rm = TRUE)
iqr_s <- q3_s - q1_s

df <- df %>%
  mutate(
    is_outlier =
      HoursStudied < q1_h - 1.5 * iqr_h |
      HoursStudied > q3_h + 1.5 * iqr_h |
      ExamScore < q1_s - 1.5 * iqr_s |
      ExamScore > q3_s + 1.5 * iqr_s
  )

df_main <- df %>% filter(!is_outlier)
df_out  <- df %>% filter(is_outlier)


#fit regression line (without outliers)
fit_no_outliers <- lm(ExamScore ~ HoursStudied, data = df_main)

slope_no_outliers <- coef(fit_no_outliers)[2]
intercept_no_outliers <- coef(fit_no_outliers)[1]
corr_no_outliers <- cor(df_main$HoursStudied, df_main$ExamScore, use = "complete.obs")
r2_no_outliers <- summary(fit_no_outliers)$r.squared

#regression with all data
fit_all <- lm(ExamScore ~ HoursStudied, data = df)

slope_all <- coef(fit_all)[2]
intercept_all <- coef(fit_all)[1]
corr_all <- cor(df$HoursStudied, df$ExamScore, use = "complete.obs")
r2_all <- summary(fit_all)$r.squared


#Hover text
df_main <- df_main %>%
  mutate(
    point_text = paste0(
      "Hours Studied: ", HoursStudied,
      "<br>Exam Score: ", ExamScore,
      "<br>Attendance: ", Attendance,
      "<br>Status: Non-outlier"
    )
  )

df_out <- df_out %>%
  mutate(
    point_text = paste0(
      "Hours Studied: ", HoursStudied,
      "<br>Exam Score: ", ExamScore,
      "<br>Attendance: ", Attendance,
      "<br>Status: Outlier"
    )
  )

#predicted x values for the regression line
x_seq <- seq(
  min(df_main$HoursStudied, na.rm = TRUE),
  max(df_main$HoursStudied, na.rm = TRUE),
  length.out = 200
)
#line without outliers
line_no_out <- data.frame(
  HoursStudied = x_seq,
  ExamScore = predict(fit_no_outliers, newdata = data.frame(HoursStudied = x_seq))
)

line_text_no_out <- paste0(
  "Trend line (no outliers)",
  "<br>Slope: ", round(slope_no_outliers, 3),
  "<br>Intercept: ", round(intercept_no_outliers, 3),
  "<br>Correlation (r): ", round(corr_no_outliers, 3),
  "<br>R²: ", round(r2_no_outliers, 3)
)

#line with all data
line_all <- data.frame(
  HoursStudied = x_seq,
  ExamScore = predict(fit_all, newdata = data.frame(HoursStudied = x_seq))
)

line_text_all <- paste0(
  "Trend line (with outliers)",
  "<br>Slope: ", round(slope_all, 3),
  "<br>Intercept: ", round(intercept_all, 3),
  "<br>Correlation (r): ", round(corr_all, 3),
  "<br>R²: ", round(r2_all, 3)
)

p <- plot_ly()

# main data: always visible
p <- p %>%
  add_markers(
    data = df_main,
    x = ~HoursStudied,
    y = ~ExamScore,
    marker = list(
      color = ~Attendance,
      coloraxis = "coloraxis",
      symbol = "circle",
      size = 8
    ),
    text = ~point_text,
    hoverinfo = "text",
    name = "Data (no outliers)",
    showlegend = FALSE
  )

# outliers: toggleable
p <- p %>%
  add_markers(
    data = df_out,
    x = ~HoursStudied,
    y = ~ExamScore,
    marker = list(
      color = ~Attendance,
      coloraxis = "coloraxis",
      symbol = "circle",
      size = 8
    ),
    text = ~point_text,
    hoverinfo = "text",
    name = "Outliers",
    legendgroup = "with_outliers",
    visible = "legendonly",
    showlegend = TRUE
  )

# trend line without outliers: always visible
p <- p %>%
  add_lines(
    data = line_no_out,
    x = ~HoursStudied,
    y = ~ExamScore,
    text = line_text_no_out,
    hoverinfo = "text",
    line = list(color = "black", width = 2),
    name = "Trend line (no outliers)",
    showlegend = FALSE
  )

# trend line with outliers: toggles with outliers
p <- p %>%
  add_lines(
    data = line_all,
    x = ~HoursStudied,
    y = ~ExamScore,
    text = line_text_all,
    hoverinfo = "text",
    line = list(color = "red", width = 2, dash = "dash"),
    name = "Trend line (with outliers)",
    legendgroup = "with_outliers",
    visible = "legendonly",
    showlegend = FALSE
  )

p %>%
  layout(
    title = "Impact of Study Time on Exam Performance and the Role of Attendance",
    xaxis = list(title = "Hours Studied"),
    yaxis = list(title = "Exam Score"),
    dragmode = "zoom",
    showlegend = TRUE,
    coloraxis = list(
      colorscale = list(
        c(0, "yellow"),
        c(1, "darkgreen")
      ),
      colorbar = list(
        title = "Attendance",
        x = 1.02,
        len = 0.75
      )
    ),
    legend = list(
      title = list(text = "Click to toggle"),
      groupclick = "togglegroup",
      x = 1.18,
      y = 0.85
    )
  ) %>%
  config(
    displayModeBar = TRUE,
    scrollZoom = TRUE
  )
```

## Rows
### Interactive Visualization 3
```{r}
df <- read.csv("StudentPerformanceFactors.csv")
p <- ggplot(df, aes(x = Sleep_Hours, y = Exam_Score)) +
  geom_jitter(alpha = 0.08, color = "#3B5998", width = 0.25, height = 0.5, size = 1) +
  geom_smooth(aes(color = "Regression Line"), method = "lm", se = TRUE, linewidth = 1.2) +
  scale_color_manual(values = "red", name = "") +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "red") +
  scale_y_continuous(limits = c(55, 85), breaks = seq(55, 85, 5)) +
  scale_x_continuous(breaks = 4:10) +
  labs(
    title = "Sleep Hours vs. Exam Score",
    subtitle = "Red line = line of best fit (linear regression). Diamonds = group averages.",
    caption = "The nearly flat regression line indicates little to no relationship between sleep hours and exam scores.",
    x = "Sleep Hours per Night",
    y = "Exam Score (0–100)"
  ) +
  theme_minimal(base_size = 13)
pl <- ggplotly(p)
for (i in seq_along(pl$x$data)) {
  nm <- pl$x$data[[i]]$name
  if (!is.null(nm) && grepl("Regression Line", nm)) {
    pl$x$data[[i]]$visible <- "legendonly"
  }
}
pl
```

### Interactive Visualization 4
```{r}
students <- read.csv("StudentPerformanceFactors.csv")
avg_study_overall <- mean(students$Hours_Studied, na.rm = TRUE)
avg_sleep_overall <- mean(students$Sleep_Hours, na.rm = TRUE)
avg_score_overall <- mean(students$Exam_Score, na.rm = TRUE)

profile_stats <- students %>%
  mutate(
    is_athlete = ifelse(Physical_Activity >= 4, "Athlete", "Non-Athlete"),
    profile_name = paste(is_athlete, ifelse(Extracurricular_Activities == "Yes", "\nWith Extracurriculars", "\nNo Extracurriculars"))
  ) %>%
  group_by(profile_name) %>%
  summarize(
    study_hrs = round(mean(Hours_Studied, na.rm = TRUE), 1),
    sleep_hrs = round(mean(Sleep_Hours, na.rm = TRUE), 1),
    exam_score = round(mean(Exam_Score, na.rm = TRUE), 1),
    .groups = 'drop'
  )

sort_study_high <- profile_stats %>% arrange(desc(study_hrs)) %>% pull(profile_name)
sort_study_low  <- profile_stats %>% arrange(study_hrs) %>% pull(profile_name)
sort_sleep_high <- profile_stats %>% arrange(desc(sleep_hrs)) %>% pull(profile_name)
sort_sleep_low  <- profile_stats %>% arrange(sleep_hrs) %>% pull(profile_name)
sort_score_high <- profile_stats %>% arrange(desc(exam_score)) %>% pull(profile_name)
sort_score_low  <- profile_stats %>% arrange(exam_score) %>% pull(profile_name)

baseline_note <- list(
  text = paste0(
    "<b>Global Averages</b><br>",
    "Study: ", round(avg_study_overall, 1), " hrs<br>",
    "Sleep: ", round(avg_sleep_overall, 1), " hrs<br>",
    "Score: ", round(avg_score_overall, 1)
  ),
  xref = "paper", yref = "paper",
  x = 0, y = 1.25,
  xanchor = "left", yanchor = "top",
  showarrow = FALSE,
  align = "left",
  bgcolor = "#f8f9fa",
  bordercolor = "#232D4B",
  borderwidth = 1,
  borderpad = 6,
  font = list(size = 11, color = "#232D4B")
)

metric_toggle_buttons <- list(
  type = "buttons",
  direction = "right",
  y = 1.25, x = 0.45,
  xanchor = "center", yanchor = "top",
  showactive = TRUE,
  buttons = list(
    list(
      method = "update",
      args = list(
        list(visible = c(TRUE, TRUE, FALSE)),
        list(yaxis = list(title = "Average Hours", gridcolor = "#e5e5e5"))
      ),
      label = "Compare Time"
    ),
    list(
      method = "update",
      args = list(
        list(visible = c(FALSE, FALSE, TRUE)),
        list(yaxis = list(title = "Average Exam Score", gridcolor = "#e5e5e5"))
      ),
      label = "View Exam Scores"
    )
  )
)

sorting_dropdown <- list(
  type = "dropdown",
  direction = "down",
  y = 1.25, x = 1,
  xanchor = "right", yanchor = "top",
  showactive = TRUE,
  buttons = list(
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_study_high)), label = "Sort: Study (High to Low)"),
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_study_low)), label = "Sort: Study (Low to High)"),
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_sleep_high)), label = "Sort: Sleep (High to Low)"),
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_sleep_low)), label = "Sort: Sleep (Low to High)"),
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_score_high)), label = "Sort: Score (High to Low)"),
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_score_low)), label = "Sort: Score (Low to High)")
  )
)

student_plot <- plot_ly(profile_stats, x = ~profile_name)

student_plot <- student_plot %>%
  add_bars(
    y = ~study_hrs,
    name = "Study Hours",
    marker = list(color = "#232D4B"),
    hovertemplate = "<b>%{x}</b><br>Study: %{y} hrs<extra></extra>",
    visible = TRUE
  ) %>%
  add_bars(
    y = ~sleep_hrs,
    name = "Sleep Hours",
    marker = list(color = "#E57200"),
    hovertemplate = "<b>%{x}</b><br>Sleep: %{y} hrs<extra></extra>",
    visible = TRUE
  ) %>%
  add_bars(
    y = ~exam_score,
    name = "Exam Score",
    marker = list(color = "#5C7F92"),
    hovertemplate = "<b>%{x}</b><br>Exam Score: %{y}<extra></extra>",
    visible = FALSE
  )

student_plot <- student_plot %>% layout(
  title = list(text = "<b>Time Allocation & Exam Performance by Lifestyle</b>", font = list(size = 18, color = "#232D4B"), y = 0.98),
  margin = list(t = 140),
  xaxis = list(title = "", tickangle = 0, tickfont = list(size = 12), categoryorder = "array", categoryarray = sort_study_high),
  yaxis = list(title = "Average Hours", gridcolor = "#e5e5e5"),
  barmode = 'group',
  paper_bgcolor = "white",
  plot_bgcolor = "white",
  annotations = list(baseline_note),
  legend = list(orientation = "h", x = 0.5, xanchor = "center", y = -0.15),
  updatemenus = list(metric_toggle_buttons, sorting_dropdown)
)

student_plot
```




