---
title: "Student Performance Factors Dashboard"
format: 
  dashboard:
    scrolling: true
    theme: cosmo
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(plotly)
library(scales)
library(DT)
library(ggplot2)

#Load data
df <- read.csv("StudentPerformanceFactors.csv")
```

## Row

### Project Overview {width=50%}
This dashboard explores the *Student Performance Factors* dataset from Kaggle, which contains information about students' academic performance and the personal and social-economical factors that may influence their exam score.
The dataset captures multiple dimensions of student experience. These include study behaviors (such as hours studied and attendance), lifestyle patterns (including sleep and physical activity), family background (such as income and parental involvement), and school-related conditions (like teacher quality and access to resources). Together, these variables provide a more holistic view of the factors that may influence academic outcomes, with **Exam_Score** as the primary measure of performance.

### Project Goals
The goal of this project is to explore how different aspects of a student’s environment and behavior are associated with exam performance, and to identify which factors appear most influential.

## Row {height=420}
### Static insight: Distribution of Exam Scores {width=50%}
```{r}
ggplot(df, aes(x = Exam_Score)) +
  geom_histogram(bins = 25, fill = "#5C7F92", color = "white") +
  labs(
    title = "Distribution of Exam Scores",
    subtitle = "Scores are roughly normally distributed, centered around 65-70.",
    x = "Exam Score",
    y = "Number of Students"
  ) +
  theme_minimal(base_size = 13)
```


### Static insight {width=50%}
```{r}
motivation_summary <- df %>%
  group_by(Motivation_Level) %>%
  summarise(avg_score = mean(Exam_Score, na.rm = TRUE), .groups = "drop")

ggplot(motivation_summary, aes(x = Motivation_Level, y = avg_score, fill = Motivation_Level)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = round(avg_score, 1)), vjust = -0.4) +
  labs(
    title = "Average Exam Score by Motivation Level",
    x = "Motivation Level",
    y = "Average Exam Score"
  ) +
  theme_minimal(base_size = 13)
```
## Row
### What these charts show
The distribution of exam scores provides a baseline view of student performance. Most students cluster within a central range, rather than at the extremes, suggesting that differences in performance are systematic rather than random. This motivates further exploration into which factors may explain this variation.
Average exam scores do not vary dramatically across motivation levels. While students with higher motivation show slightly better performance, the differences are relatively small. This suggests that motivation alone may not be a strong explanatory factor, and that other variables—such as study habits or external support—may play a more significant role.

More specifically, this dashboard aims to:

- Examine whether **academic habits** (such as hours studied and attendance) are strong predictors of performance  
- Investigate the impact of **support systems**, including parental involvement and access to resources  
- research question 3
- research question 4

By combining multiple perspectives, the dashboard seeks to move beyond simple explanations of success and highlight the complexity of student achievement.

## Row 
### Interactive Visualization 1

```{r}
df$Family_Income        <- as.character(df$Family_Income)
df$Parental_Involvement <- as.character(df$Parental_Involvement)
df$Access_to_Resources  <- as.character(df$Access_to_Resources)

pi_levels  <- c("High", "Medium", "Low")
atr_levels <- c("High", "Medium", "Low")
pi_opts    <- c("All", pi_levels)
atr_opts   <- c("All", atr_levels)

income_colors <- c(
  "Low"    = "#EF553B",
  "Medium" = "#00CC96",
  "High"   = "#636EFA"
)

combos <- expand.grid(pi = pi_opts, atr = atr_opts, stringsAsFactors = FALSE)

fig            <- plot_ly()
visibility_map <- list()
trace_index    <- 0

for (i in seq_len(nrow(combos))) {
  pi_val     <- combos$pi[i]
  atr_val    <- combos$atr[i]
  is_default <- (pi_val == "All" && atr_val == "All")

  sub <- df
  if (pi_val  != "All") sub <- sub[sub$Parental_Involvement == pi_val, ]
  if (atr_val != "All") sub <- sub[sub$Access_to_Resources  == atr_val, ]

  sub$Family_Income <- factor(sub$Family_Income, levels = c("Low", "Medium", "High"))

  for (inc in c("Low", "Medium", "High")) {
    sub_inc <- sub[sub$Family_Income == inc, ]
    scores  <- sub_inc$Exam_Score

    # Pre-compute box stats
    mn  <- round(mean(scores, na.rm = TRUE), 2)
    q1  <- quantile(scores, 0.25, na.rm = TRUE)
    med <- median(scores, na.rm = TRUE)
    q3  <- quantile(scores, 0.75, na.rm = TRUE)
    iqr <- q3 - q1
    lo  <- round(min(scores, na.rm = TRUE), 2)
    hi  <- round(max(scores, na.rm = TRUE), 2)

    hover_text <- paste0(
      "<b>", inc, " Income</b><br>",
      "Max: ", hi,  "<br>",
      "Mean: ", mn, "<br>",
      "Median: ", med, "<br>",
      "Min: ", lo
    )

    fig <- fig %>% add_trace(
      x              = sub_inc$Family_Income,
      y              = scores,
      type           = "box",
      name           = inc,
      marker         = list(color = income_colors[[inc]]),
      line           = list(color = income_colors[[inc]]),
      fillcolor      = paste0(income_colors[[inc]], "55"),
      boxpoints      = "outliers",
      hoverinfo      = "none",          # disable box hover entirely
      visible        = is_default,
      legendgroup    = inc,
      showlegend     = (i == 0),
      offsetgroup    = inc,
      alignmentgroup = "fixed"
    )

    trace_index <- trace_index + 1
    visibility_map[[length(visibility_map) + 1]] <- list(
      pi = pi_val, atr = atr_val, trace = trace_index
    )

    # Invisible scatter point at median — carries the unified tooltip
    fig <- fig %>% add_trace(
      x          = inc,
      y          = med,
      type       = "scatter",
      mode       = "markers",
      name       = inc,
      legendgroup = inc,
      showlegend = FALSE,
      marker     = list(color = "rgba(0,0,0,0)", size = 40),  # fully transparent, large hit area
      hovertemplate = paste0(
        "<b>", inc, " Income</b><br>",
        "Max: ", hi,  "<br>",
        "Mean: ", mn, "<br>",
        "Median: ", med, "<br>",
        "Min: ", lo,
        "<extra></extra>"
      ),
      visible    = is_default
    )

    trace_index <- trace_index + 1
    visibility_map[[length(visibility_map) + 1]] <- list(
      pi = pi_val, atr = atr_val, trace = trace_index
    )
    
    # Outlier points with tooltips
    outliers <- scores[scores < q1 - 1.5 * iqr | scores > q3 + 1.5 * iqr]
    if (length(outliers) > 0) {
      fig <- fig %>% add_trace(
        x             = rep(inc, length(outliers)),
        y             = outliers,
        type          = "scatter",
        mode          = "markers",
        name          = inc,
        legendgroup   = inc,
        showlegend    = FALSE,
        marker        = list(color = income_colors[[inc]], size = 6),
        hoverlabel    = list(bgcolor = "#000000", font = list(color = "#ffffff")),
        hovertemplate = paste0(
          "<b>", inc, " Income</b><br>",
          "Outlier: %{y}<br>",
          "<extra></extra>"
        ),
        visible = is_default
      )

      trace_index <- trace_index + 1
      visibility_map[[length(visibility_map) + 1]] <- list(
        pi = pi_val, atr = atr_val, trace = trace_index
      )
    }
  }
}

make_vis <- function(sel_pi, sel_atr) {
  vapply(visibility_map, function(m) m$pi == sel_pi && m$atr == sel_atr, logical(1))
}

pi_buttons <- lapply(pi_opts, function(pv) {
  list(method = "update",
       args   = list(list(visible = as.list(make_vis(pv, "All")))),
       label  = pv)
})

atr_buttons <- lapply(atr_opts, function(av) {
  list(method = "update",
       args   = list(list(visible = as.list(make_vis("All", av)))),
       label  = av)
})

fig <- fig %>%
  layout(
    title = list(
      text    = "Effect of Family Income on Exam Performance",
      x       = 0.5,
      xanchor = "center",
      y       = 0.97,
      yanchor = "top",
      font    = list(size = 16)
    ),
    xaxis = list(
      title         = "Family Income",
      type          = "category",
      categoryorder = "array",
      categoryarray = c("Low", "Medium", "High")
    ),
    yaxis   = list(title = "Exam Score"),
    boxmode = "overlay",
    margin  = list(t = 130, r = 30, b = 60, l = 60),

    updatemenus = list(
      list(
        type        = "dropdown",
        active      = 0,
        x           = 0.01,
        xanchor     = "left",
        y           = 1.10,
        yanchor     = "top",
        buttons     = pi_buttons,
        bgcolor     = "#ffffff",
        bordercolor = "#aaaaaa",
        font        = list(size = 12)
      ),
      list(
        type        = "dropdown",
        active      = 0,
        x           = 0.13,
        xanchor     = "left",
        y           = 1.10,
        yanchor     = "top",
        buttons     = atr_buttons,
        bgcolor     = "#ffffff",
        bordercolor = "#aaaaaa",
        font        = list(size = 12)
      )
    ),

    annotations = list(
      list(
        text      = "<b>Parental Involvement:</b>",
        x = 0,  xref = "paper", xanchor = "left",
        y = 1.17, yref = "paper", yanchor = "top",
        showarrow = FALSE, font = list(size = 12, color = "#333333")
      ),
      list(
        text      = "<b>Access to Resources:</b>",
        x = 0.12, xref = "paper", xanchor = "left",
        y = 1.17, yref = "paper", yanchor = "top",
        showarrow = FALSE, font = list(size = 12, color = "#333333")
      )
    )
  )

fig
```

### Interactive visualization 2

```{r}
df <- read.csv("StudentPerformanceFactors.csv")
#rename columns
df <- df %>%
  rename(
     HoursStudied = Hours_Studied,
     ExamScore = Exam_Score,
     Attendance = Attendance
   )

#filter outliers
q1_h <- quantile(df$HoursStudied, 0.25, na.rm = TRUE)
q3_h <- quantile(df$HoursStudied, 0.75, na.rm = TRUE)
iqr_h <- q3_h - q1_h

q1_s <- quantile(df$ExamScore, 0.25, na.rm = TRUE)
q3_s <- quantile(df$ExamScore, 0.75, na.rm = TRUE)
iqr_s <- q3_s - q1_s

df <- df %>%
  mutate(
    is_outlier =
      HoursStudied < q1_h - 1.5 * iqr_h |
      HoursStudied > q3_h + 1.5 * iqr_h |
      ExamScore < q1_s - 1.5 * iqr_s |
      ExamScore > q3_s + 1.5 * iqr_s
  )

df_main <- df %>% filter(!is_outlier)
df_out  <- df %>% filter(is_outlier)


#fit regression line (without outliers)
fit_no_outliers <- lm(ExamScore ~ HoursStudied, data = df_main)

slope_no_outliers <- coef(fit_no_outliers)[2]
intercept_no_outliers <- coef(fit_no_outliers)[1]
corr_no_outliers <- cor(df_main$HoursStudied, df_main$ExamScore, use = "complete.obs")
r2_no_outliers <- summary(fit_no_outliers)$r.squared

#regression with all data
fit_all <- lm(ExamScore ~ HoursStudied, data = df)

slope_all <- coef(fit_all)[2]
intercept_all <- coef(fit_all)[1]
corr_all <- cor(df$HoursStudied, df$ExamScore, use = "complete.obs")
r2_all <- summary(fit_all)$r.squared


#Hover text
df_main <- df_main %>%
  mutate(
    point_text = paste0(
      "Hours Studied: ", HoursStudied,
      "<br>Exam Score: ", ExamScore,
      "<br>Attendance: ", Attendance,
      "<br>Status: Non-outlier"
    )
  )

df_out <- df_out %>%
  mutate(
    point_text = paste0(
      "Hours Studied: ", HoursStudied,
      "<br>Exam Score: ", ExamScore,
      "<br>Attendance: ", Attendance,
      "<br>Status: Outlier"
    )
  )

#predicted x values for the regression line
x_seq <- seq(
  min(df_main$HoursStudied, na.rm = TRUE),
  max(df_main$HoursStudied, na.rm = TRUE),
  length.out = 200
)
#line without outliers
line_no_out <- data.frame(
  HoursStudied = x_seq,
  ExamScore = predict(fit_no_outliers, newdata = data.frame(HoursStudied = x_seq))
)

line_text_no_out <- paste0(
  "Trend line (no outliers)",
  "<br>Slope: ", round(slope_no_outliers, 3),
  "<br>Intercept: ", round(intercept_no_outliers, 3),
  "<br>Correlation (r): ", round(corr_no_outliers, 3),
  "<br>R²: ", round(r2_no_outliers, 3)
)

#line with all data
line_all <- data.frame(
  HoursStudied = x_seq,
  ExamScore = predict(fit_all, newdata = data.frame(HoursStudied = x_seq))
)

line_text_all <- paste0(
  "Trend line (with outliers)",
  "<br>Slope: ", round(slope_all, 3),
  "<br>Intercept: ", round(intercept_all, 3),
  "<br>Correlation (r): ", round(corr_all, 3),
  "<br>R²: ", round(r2_all, 3)
)

p <- plot_ly()

# main data: always visible
p <- p %>%
  add_markers(
    data = df_main,
    x = ~HoursStudied,
    y = ~ExamScore,
    marker = list(
      color = ~Attendance,
      coloraxis = "coloraxis",
      symbol = "circle",
      size = 8
    ),
    text = ~point_text,
    hoverinfo = "text",
    name = "Data (no outliers)",
    showlegend = FALSE
  )

# outliers: toggleable
p <- p %>%
  add_markers(
    data = df_out,
    x = ~HoursStudied,
    y = ~ExamScore,
    marker = list(
      color = ~Attendance,
      coloraxis = "coloraxis",
      symbol = "circle",
      size = 8
    ),
    text = ~point_text,
    hoverinfo = "text",
    name = "Outliers",
    legendgroup = "with_outliers",
    visible = "legendonly",
    showlegend = TRUE
  )

# trend line without outliers: always visible
p <- p %>%
  add_lines(
    data = line_no_out,
    x = ~HoursStudied,
    y = ~ExamScore,
    text = line_text_no_out,
    hoverinfo = "text",
    line = list(color = "black", width = 2),
    name = "Trend line (no outliers)",
    showlegend = FALSE
  )

# trend line with outliers: toggles with outliers
p <- p %>%
  add_lines(
    data = line_all,
    x = ~HoursStudied,
    y = ~ExamScore,
    text = line_text_all,
    hoverinfo = "text",
    line = list(color = "red", width = 2, dash = "dash"),
    name = "Trend line (with outliers)",
    legendgroup = "with_outliers",
    visible = "legendonly",
    showlegend = FALSE
  )

p %>%
  layout(
    title = "Impact of Study Time on Exam Performance and the Role of Attendance",
    xaxis = list(title = "Hours Studied"),
    yaxis = list(title = "Exam Score"),
    dragmode = "zoom",
    showlegend = TRUE,
    coloraxis = list(
      colorscale = list(
        c(0, "yellow"),
        c(1, "darkgreen")
      ),
      colorbar = list(
        title = "Attendance",
        x = 1.02,
        len = 0.75
      )
    ),
    legend = list(
      title = list(text = "Click to toggle"),
      groupclick = "togglegroup",
      x = 1.18,
      y = 0.85
    )
  ) %>%
  config(
    displayModeBar = TRUE,
    scrollZoom = TRUE
  )
```

## Rows
### Interactive Visualization 3
```{r}
df <- read.csv("StudentPerformanceFactors.csv")
p <- ggplot(df, aes(x = Sleep_Hours, y = Exam_Score)) +
  geom_jitter(alpha = 0.08, color = "#3B5998", width = 0.25, height = 0.5, size = 1) +
  geom_smooth(aes(color = "Regression Line"), method = "lm", se = TRUE, linewidth = 1.2) +
  scale_color_manual(values = "red", name = "") +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "red") +
  scale_y_continuous(limits = c(55, 85), breaks = seq(55, 85, 5)) +
  scale_x_continuous(breaks = 4:10) +
  labs(
    title = "Sleep Hours vs. Exam Score",
    subtitle = "Red line = line of best fit (linear regression). Diamonds = group averages.",
    caption = "The nearly flat regression line indicates little to no relationship between sleep hours and exam scores.",
    x = "Sleep Hours per Night",
    y = "Exam Score (0–100)"
  ) +
  theme_minimal(base_size = 13)
pl <- ggplotly(p)
for (i in seq_along(pl$x$data)) {
  nm <- pl$x$data[[i]]$name
  if (!is.null(nm) && grepl("Regression Line", nm)) {
    pl$x$data[[i]]$visible <- "legendonly"
  }
}
pl
```

### Interactive Visualization 4
```{r}
library(dplyr)
library(plotly)

students <- read.csv("StudentPerformanceFactors.csv")

profile_stats <- students %>%
  mutate(
    is_athlete = ifelse(Physical_Activity >= 4, "Athlete", "Non-Athlete"),
    has_extra = ifelse(Extracurricular_Activities == "Yes", "+ Extras", "No Extras")
  ) %>%
  group_by(is_athlete, has_extra) %>%
  summarize(
    study_hrs = round(mean(Hours_Studied, na.rm = TRUE), 1),
    sleep_hrs = round(mean(Sleep_Hours, na.rm = TRUE), 1),
    exam_score = round(mean(Exam_Score, na.rm = TRUE), 1),
    .groups = 'drop'
  ) %>%
  mutate(
    profile_name = paste0(is_athlete, "<br>", has_extra)
  )

gap_row <- data.frame(
  is_athlete = "Gap",
  has_extra = "Gap",
  profile_name = " ", 
  study_hrs = NA,
  sleep_hrs = NA,
  exam_score = NA
)

plot_data <- bind_rows(
  profile_stats %>% filter(is_athlete == "Athlete"),
  gap_row,
  profile_stats %>% filter(is_athlete == "Non-Athlete")
)

ath <- profile_stats %>% filter(is_athlete == "Athlete")
non <- profile_stats %>% filter(is_athlete == "Non-Athlete")

sort_study_desc <- c(ath %>% arrange(desc(study_hrs)) %>% pull(profile_name), " ", non %>% arrange(desc(study_hrs)) %>% pull(profile_name))
sort_study_asc  <- c(ath %>% arrange(study_hrs) %>% pull(profile_name), " ", non %>% arrange(study_hrs) %>% pull(profile_name))
sort_sleep_desc <- c(ath %>% arrange(desc(sleep_hrs)) %>% pull(profile_name), " ", non %>% arrange(desc(sleep_hrs)) %>% pull(profile_name))
sort_sleep_asc  <- c(ath %>% arrange(sleep_hrs) %>% pull(profile_name), " ", non %>% arrange(sleep_hrs) %>% pull(profile_name))
sort_score_desc <- c(ath %>% arrange(desc(exam_score)) %>% pull(profile_name), " ", non %>% arrange(desc(exam_score)) %>% pull(profile_name))
sort_score_asc  <- c(ath %>% arrange(exam_score) %>% pull(profile_name), " ", non %>% arrange(exam_score) %>% pull(profile_name))

plot_data$profile_name <- factor(plot_data$profile_name, levels = sort_study_desc)

time_plot <- plot_ly(plot_data, x = ~profile_name) %>%
  add_bars(
    y = ~study_hrs,
    name = "Study Hours",
    marker = list(color = "#232D4B"),
    hovertemplate = "<b>%{x}</b><br>Study: %{y} hrs<extra></extra>"
  ) %>%
  add_bars(
    y = ~sleep_hrs,
    name = "Sleep Hours",
    marker = list(color = "#E57200"),
    hovertemplate = "<b>%{x}</b><br>Sleep: %{y} hrs<extra></extra>"
  ) %>%
  layout(
    yaxis = list(title = "Average Hours", gridcolor = "#e5e5e5", range = c(0, 22)),
    barmode = 'group'
  )

score_plot <- plot_ly(plot_data, x = ~profile_name) %>%
  add_bars(
    y = ~exam_score,
    name = "Final Exam Score",
    marker = list(color = "#5C7F92"),
    hovertemplate = "<b>%{x}</b><br>Exam Score: %{y}<extra></extra>"
  ) %>%
  layout(
    yaxis = list(title = "Average Score", gridcolor = "#e5e5e5", range = c(0, 110))
  )

sorting_dropdown <- list(
  type = "dropdown",
  direction = "down",
  x = 1,
  xref = "container",
  xanchor = "right",
  y = 0.98,
  yref = "container",
  yanchor = "top",
  buttons = list(
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_study_desc, "xaxis2.categoryorder" = "array", "xaxis2.categoryarray" = sort_study_desc)), label = "Sort: Study (High to Low)"),
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_study_asc, "xaxis2.categoryorder" = "array", "xaxis2.categoryarray" = sort_study_asc)), label = "Sort: Study (Low to High)"),
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_sleep_desc, "xaxis2.categoryorder" = "array", "xaxis2.categoryarray" = sort_sleep_desc)), label = "Sort: Sleep (High to Low)"),
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_sleep_asc, "xaxis2.categoryorder" = "array", "xaxis2.categoryarray" = sort_sleep_asc)), label = "Sort: Sleep (Low to High)"),
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_score_desc, "xaxis2.categoryorder" = "array", "xaxis2.categoryarray" = sort_score_desc)), label = "Sort: Score (High to Low)"),
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_score_asc, "xaxis2.categoryorder" = "array", "xaxis2.categoryarray" = sort_score_asc)), label = "Sort: Score (Low to High)")
  )
)

final_plot <- subplot(time_plot, score_plot, nrows = 1, margin = 0.05, titleX = FALSE, titleY = TRUE) %>%
  layout(
    title = list(
      text = "<b>Impact of Athletics and Extracurriculars on Habits and Performance</b>",
      font = list(size = 18, color = "#232D4B"),
      x = 0.5,
      xanchor = "center",
      y = 0.98,
      yref = "container"
    ),
    annotations = list(
      list(x = 0.24, y = 1.02, text = "<b>Study and Sleep Time</b>", showarrow = FALSE, xref = "paper", yref = "paper", font = list(size = 14), xanchor = "center", yanchor = "bottom"),
      list(x = 0.76, y = 1.02, text = "<b>Exam Scores</b>", showarrow = FALSE, xref = "paper", yref = "paper", font = list(size = 14), xanchor = "center", yanchor = "bottom")
    ),
    legend = list(orientation = "h", x = 0.5, xanchor = "center", y = -0.35),
    margin = list(t = 120, b = 120),
    paper_bgcolor = "white",
    plot_bgcolor = "white",
    barmode = 'group',
    xaxis = list(tickangle = -45, tickfont = list(size = 11)),
    xaxis2 = list(tickangle = -45, tickfont = list(size = 11)),
    updatemenus = list(sorting_dropdown)
  )

final_plot
```




