---
title: "Student Performance Factors Dashboard"
format: 
  dashboard:
    scrolling: true
    theme: flatly
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
css: styles.css
---

```{r setup, include=FALSE}
library(tidyverse)
library(plotly)
library(scales)
library(DT)
library(ggplot2)

#Load data
df <- read.csv("StudentPerformanceFactors.csv")

# color palette
color_high <- "#18BC9C"
color_med  <- "#F39C12"
color_low  <- "#E74C3C"
color_base <- "#2C3E50"
```

# Overview 
## Row {height= 180}
::: {.card title="Project Overview" fill="false"}
This dashboard explores the *Student Performance Factors* dataset from Kaggle.

It examines how different aspects of a student’s life influence their **Exam Score**, including:
- **Study behavior** → hours studied, attendance  
- **Lifestyle** → sleep, physical activity  
- **Family background** → income, parental involvement  
- **School environment** → teacher quality, access to resources  

Together, these factors provide a **holistic view of academic performance**.
:::

::: {.card title="Research Questions" fill="false"}
More specifically, this dashboard aims to:

- Investigate the impact of **support systems**, including parental involvement and access to resources  

- Examine whether **academic habits** (such as hours studied and attendance) are strong predictors of performance  

- Explore how **lifestyle habits** (such as sleep duration) differentiate the highest and lowest-performing students

- Assess how **extracurricular involvement** (like athletics and clubs) impacts a student's allocation of sleep and study time
:::


## Row
### Static insight: Distribution of Exam Scores {width=50%}
```{r}
ggplot(df, aes(x = Exam_Score)) +
  geom_histogram(bins = 25, fill = color_base, color = "white") +
  labs(
    title = "Distribution of Exam Scores",
    subtitle = "Scores are roughly normally distributed, centered around 65-70.",
    x = "Exam Score",
    y = "Number of Students"
  ) +
  theme_minimal(base_size = 13)
```


### Static insight {width=50%}
```{r}
motivation_summary <- df %>%
  group_by(Motivation_Level) %>%
  summarise(avg_score = mean(Exam_Score, na.rm = TRUE), .groups = "drop")

motivation_summary$Motivation_Level <- factor(
  motivation_summary$Motivation_Level,
  levels = c("Low", "Medium", "High")
)

ggplot(motivation_summary, aes(x = Motivation_Level, y = avg_score, fill = Motivation_Level)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = round(avg_score, 1)), vjust = -0.4) +
  scale_fill_manual(values = c(
    "Low" = color_low,
    "Medium" = color_med,
    "High" = color_high
  )) +
  labs(
    title = "Average Exam Score by Motivation Level",
    x = "Motivation Level",
    y = "Average Exam Score"
  ) +
  theme_minimal(base_size = 13)
```


## Row {height=220}
::: {.card title="What the Distribution Shows"}
The distribution of exam scores provides a baseline view of student performance. 

Most students cluster around a **central score range**, with few extreme values.
→ Performance differences appear **systematic, not random** 
→ Suggests other factors are influencing outcomes, this motivates further exploration.
:::


::: {.card title="What the Motivation Chart Shows"}
Average exam scores do not vary dramatically across motivation levels. 

Higher motivation is linked to **slightly better scores**, but the effect is small.
→ Motivation alone is **not a strong predictor**  
→ Other factors (study habits, support) likely matter more
:::

## Row {height=220}
### Goals

::: {.card title="Project Goals" fill="false}
**Goal:** Understand what drives student performance

- Analyze how **behavior, environment, and lifestyle** relate to exam scores  
- Identify which factors are **most influential**  
- Show that success is **multi-dimensional**, not explained by one variable  

<a href="#interactive-analysis" class="floating-next">Go to Interactive Analysis →</a>
:::

# Interactive Analysis
::: {.card title="Dashboard Navigation Guide" fill="false"}

Academic success isn't driven by a single metric. To uncover what truly impacts exam scores, this dashboard is structured to guide you through four distinct layers of a student's life:

1. **Foundation (Family & Support):** Does economic advantage matter, or can parental involvement bridge the gap?
2. **Academic Input (Study & Attendance):** How do raw study hours translate to grades when class attendance fluctuates?
3. **Recovery (Sleep & Proximity):** Is sleep the ultimate performance booster, or does commute time interfere?
4. **Time Management (Athletics & Clubs):** Do heavily involved students suffer academically, or are they better time managers?

Follow the charts below to explore how these layers interact. View the "Key Insights" on the right side of each chart for guidance on interactivity and takeaways.
:::

## Row {height=500}

### Chart {width=70%}

```{r}
df$Family_Income        <- as.character(df$Family_Income)
df$Parental_Involvement <- as.character(df$Parental_Involvement)
df$Access_to_Resources  <- as.character(df$Access_to_Resources)

pi_levels  <- c("High", "Medium", "Low")
atr_levels <- c("High", "Medium", "Low")
pi_opts    <- c("All", pi_levels)
atr_opts   <- c("All", atr_levels)

income_colors <- c(
  "Low" = color_low, 
  "Medium" = color_med, 
  "High" = color_high
)

combos <- expand.grid(pi = pi_opts, atr = atr_opts, stringsAsFactors = FALSE)

fig            <- plot_ly()
visibility_map <- list()
trace_index    <- 0

for (i in seq_len(nrow(combos))) {
  pi_val     <- combos$pi[i]
  atr_val    <- combos$atr[i]
  is_default <- (pi_val == "All" && atr_val == "All")

  sub <- df
  if (pi_val  != "All") sub <- sub[sub$Parental_Involvement == pi_val, ]
  if (atr_val != "All") sub <- sub[sub$Access_to_Resources  == atr_val, ]

  sub$Family_Income <- factor(sub$Family_Income, levels = c("Low", "Medium", "High"))

  for (inc in c("Low", "Medium", "High")) {
    sub_inc <- sub[sub$Family_Income == inc, ]
    scores  <- sub_inc$Exam_Score

    mn  <- round(mean(scores, na.rm = TRUE), 2)
    q1  <- quantile(scores, 0.25, na.rm = TRUE)
    med <- median(scores, na.rm = TRUE)
    q3  <- quantile(scores, 0.75, na.rm = TRUE)
    iqr <- q3 - q1
    lo  <- round(min(scores, na.rm = TRUE), 2)
    hi  <- round(max(scores, na.rm = TRUE), 2)

    hover_text <- paste0(
      "<b>", inc, " Income</b><br>",
      "Max: ", hi,  "<br>",
      "Mean: ", mn, "<br>",
      "Median: ", med, "<br>",
      "Min: ", lo
    )

    fig <- fig %>% add_trace(
      x              = sub_inc$Family_Income,
      y              = scores,
      type           = "box",
      name           = inc,
      marker         = list(color = income_colors[[inc]]),
      line           = list(color = income_colors[[inc]]),
      fillcolor      = paste0(income_colors[[inc]], "55"),
      boxpoints      = "outliers",
      hoverinfo      = "none",
      visible        = is_default,
      legendgroup    = inc,
      showlegend     = (i == 0),
      offsetgroup    = inc,
      alignmentgroup = "fixed"
    )

    trace_index <- trace_index + 1
    visibility_map[[length(visibility_map) + 1]] <- list(
      pi = pi_val, atr = atr_val, trace = trace_index
    )

    fig <- fig %>% add_trace(
      x          = inc,
      y          = med,
      type       = "scatter",
      mode       = "markers",
      name       = inc,
      legendgroup = inc,
      showlegend = FALSE,
      marker     = list(color = "rgba(0,0,0,0)", size = 40),
      hovertemplate = paste0(
        "<b>", inc, " Income</b><br>",
        "Max: ", hi,  "<br>",
        "Mean: ", mn, "<br>",
        "Median: ", med, "<br>",
        "Min: ", lo,
        "<extra></extra>"
      ),
      visible    = is_default
    )

    trace_index <- trace_index + 1
    visibility_map[[length(visibility_map) + 1]] <- list(
      pi = pi_val, atr = atr_val, trace = trace_index
    )
    
    outliers <- scores[scores < q1 - 1.5 * iqr | scores > q3 + 1.5 * iqr]
    if (length(outliers) > 0) {
      fig <- fig %>% add_trace(
        x             = rep(inc, length(outliers)),
        y             = outliers,
        type          = "scatter",
        mode          = "markers",
        name          = inc,
        legendgroup   = inc,
        showlegend    = FALSE,
        marker        = list(color = income_colors[[inc]], size = 6),
        hoverlabel    = list(bgcolor = "#000000", font = list(color = "#ffffff")),
        hovertemplate = paste0(
          "<b>", inc, " Income</b><br>",
          "Outlier: %{y}<br>",
          "<extra></extra>"
        ),
        visible = is_default
      )

      trace_index <- trace_index + 1
      visibility_map[[length(visibility_map) + 1]] <- list(
        pi = pi_val, atr = atr_val, trace = trace_index
      )
    }
  }
}

make_vis <- function(sel_pi, sel_atr) {
  vapply(visibility_map, function(m) m$pi == sel_pi && m$atr == sel_atr, logical(1))
}

pi_buttons <- lapply(pi_opts, function(pv) {
  list(method = "update",
       args   = list(list(visible = as.list(make_vis(pv, "All")))),
       label  = pv)
})

atr_buttons <- lapply(atr_opts, function(av) {
  list(method = "update",
       args   = list(list(visible = as.list(make_vis("All", av)))),
       label  = av)
})

fig <- fig %>%
  layout(
    title = list(
      text    = "Effect of Family Income on Exam Performance",
      x       = 0.5,
      xanchor = "center",
      y       = 0.97,
      yanchor = "top",
      font    = list(size = 16)
    ),
    xaxis = list(
      title         = "Family Income",
      type          = "category",
      categoryorder = "array",
      categoryarray = c("Low", "Medium", "High")
    ),
    yaxis   = list(title = "Exam Score"),
    boxmode = "overlay",
    margin  = list(t = 130, r = 30, b = 60, l = 60),

    updatemenus = list(
      list(
        type        = "dropdown",
        active      = 0,
        x           = 0.01,
        xanchor     = "left",
        y           = 1.10,
        yanchor     = "top",
        buttons     = pi_buttons,
        bgcolor     = "#ffffff",
        bordercolor = "#aaaaaa",
        font        = list(size = 12)
      ),
      list(
        type        = "dropdown",
        active      = 0,
        x           = 0.25,
        xanchor     = "left",
        y           = 1.10,
        yanchor     = "top",
        buttons     = atr_buttons,
        bgcolor     = "#ffffff",
        bordercolor = "#aaaaaa",
        font        = list(size = 12)
      )
    ),

    annotations = list(
      list(
        text      = "<b>Parental Involvement:</b>",
        x = 0,  xref = "paper", xanchor = "left",
        y = 1.17, yref = "paper", yanchor = "top",
        showarrow = FALSE, font = list(size = 12, color = "#333333")
      ),
      list(
        text      = "<b>Access to Resources:</b>",
        x = 0.24, xref = "paper", xanchor = "left",
        y = 1.17, yref = "paper", yanchor = "top",
        showarrow = FALSE, font = list(size = 12, color = "#333333")
      )
    )
  )

fig
```

### Key Insights {width=30%}

::: {.card title="Key Insights"}
**What This Shows**

Exam score distributions broken down by family income level, with filters to layer in parental involvement and access to resources.

**Interactive Features**

- Use the **Parental Involvement** and **Access to Resources** dropdowns to explore how these support factors interact with income.
- Hover over each box to see the median, mean, min, and max for that income group.
- Hover over individual points to identify outlier scores.
- Hold down mouse and drag to zoom in.

**Family Income & Performance**

- Higher family income is associated with slightly higher median exam scores, but distributions overlap considerably across all three groups. Income alone does not determine exam performance
- Look at the width and position of each box: the similarity in spread suggests the relationship is more nuanced than a simple income advantage.

**The Role of Support Systems**

- When both parental involvement and access to resources are high, the performance gap between income groups narrows noticeably.
- Comparing low-income students with high parental involvement against high-income students with low parental involvement reveals a clear advantage for the former, showing that involvement can actively offset economic disadvantage.

**Key Takeaway**

Income correlates with slightly higher scores, but it isn't the only factor. Support systems, like parental involvement and access to resources, can help reduce the performance gap, suggesting that targeted support may be as impactful as income itself.
:::

## Row {height=500}

### Chart {width=70%}

```{r}
df <- read.csv("StudentPerformanceFactors.csv")

# rename columns
df <- df %>%
  rename(
    HoursStudied = Hours_Studied,
    ExamScore = Exam_Score,
    Attendance = Attendance
  )

# filter outliers
q1_h <- quantile(df$HoursStudied, 0.25, na.rm = TRUE)
q3_h <- quantile(df$HoursStudied, 0.75, na.rm = TRUE)
iqr_h <- q3_h - q1_h

q1_s <- quantile(df$ExamScore, 0.25, na.rm = TRUE)
q3_s <- quantile(df$ExamScore, 0.75, na.rm = TRUE)
iqr_s <- q3_s - q1_s

df <- df %>%
  mutate(
    is_outlier =
      HoursStudied < q1_h - 1.5 * iqr_h |
      HoursStudied > q3_h + 1.5 * iqr_h |
      ExamScore < q1_s - 1.5 * iqr_s |
      ExamScore > q3_s + 1.5 * iqr_s
  )

df_main <- df %>% filter(!is_outlier)
df_out  <- df %>% filter(is_outlier)

# fit regression line (without outliers)
fit_no_outliers <- lm(ExamScore ~ HoursStudied, data = df_main)

slope_no_outliers <- coef(fit_no_outliers)[2]
intercept_no_outliers <- coef(fit_no_outliers)[1]
corr_no_outliers <- cor(df_main$HoursStudied, df_main$ExamScore, use = "complete.obs")
r2_no_outliers <- summary(fit_no_outliers)$r.squared

# regression with all data
fit_all <- lm(ExamScore ~ HoursStudied, data = df)

slope_all <- coef(fit_all)[2]
intercept_all <- coef(fit_all)[1]
corr_all <- cor(df$HoursStudied, df$ExamScore, use = "complete.obs")
r2_all <- summary(fit_all)$r.squared

# hover text
df_main <- df_main %>%
  mutate(
    point_text = paste0(
      "Hours Studied: ", HoursStudied,
      "<br>Exam Score: ", ExamScore,
      "<br>Attendance: ", Attendance,
      "<br>Status: Non-outlier"
    )
  )

df_out <- df_out %>%
  mutate(
    point_text = paste0(
      "Hours Studied: ", HoursStudied,
      "<br>Exam Score: ", ExamScore,
      "<br>Attendance: ", Attendance,
      "<br>Status: Outlier"
    )
  )

# predicted x values for regression lines
x_seq <- seq(
  min(df$HoursStudied, na.rm = TRUE),
  max(df$HoursStudied, na.rm = TRUE),
  length.out = 200
)

# line without outliers
line_no_out <- data.frame(
  HoursStudied = x_seq,
  ExamScore = predict(fit_no_outliers, newdata = data.frame(HoursStudied = x_seq))
)

line_text_no_out <- paste0(
  "Trend line (no outliers)",
  "<br>Slope: ", round(slope_no_outliers, 3),
  "<br>Intercept: ", round(intercept_no_outliers, 3),
  "<br>Correlation (r): ", round(corr_no_outliers, 3),
  "<br>R²: ", round(r2_no_outliers, 3)
)

# line with all data
line_all <- data.frame(
  HoursStudied = x_seq,
  ExamScore = predict(fit_all, newdata = data.frame(HoursStudied = x_seq))
)

line_text_all <- paste0(
  "Trend line (with outliers)",
  "<br>Slope: ", round(slope_all, 3),
  "<br>Intercept: ", round(intercept_all, 3),
  "<br>Correlation (r): ", round(corr_all, 3),
  "<br>R²: ", round(r2_all, 3)
)

p <- plot_ly()


# main data, green, visible by default
p <- p %>%
  add_markers(
    data = df_main,
    x = ~HoursStudied,
    y = ~ExamScore,
    marker = list(
      color = color_high,
      symbol = "circle",
      size = 8
    ),
    text = ~point_text,
    hoverinfo = "text",
    name = "Students",
    legendgroup = "plain_points",
    showlegend = FALSE,
    visible = TRUE
  )

# outliers, green, hidden by default
p <- p %>%
  add_markers(
    data = df_out,
    x = ~HoursStudied,
    y = ~ExamScore,
    marker = list(
      color = color_high,
      symbol = "circle",
      size = 8
    ),
    text = ~point_text,
    hoverinfo = "text",
    name = "Outliers",
    legendgroup = "with_outliers_plain",
    visible = "legendonly",
    showlegend = TRUE
  )


# main data colored by attendance, hidden by default
p <- p %>%
  add_markers(
    data = df_main,
    x = ~HoursStudied,
    y = ~ExamScore,
    marker = list(
  color = ~Attendance,
  colorscale = list(
    c(0, color_low),
    c(1, color_high)
  ),
  cmin = min(df$Attendance, na.rm = TRUE),
  cmax = max(df$Attendance, na.rm = TRUE),
  showscale = TRUE,
  colorbar = list(
    title = "Attendance",
    x = 1.02,
    len = 0.75
  ),
  symbol = "circle",
  size = 8
),
    text = ~point_text,
    hoverinfo = "text",
    name = "Color by Attendance",
    legendgroup = "attendance_colored",
    visible = "legendonly",
    showlegend = TRUE
  )

# outliers colored by attendance, hidden by default
p <- p %>%
  add_markers(
    data = df_out,
    x = ~HoursStudied,
    y = ~ExamScore,
    marker = list(
  color = ~Attendance,
  colorscale = list(
    c(0, color_high),
    c(1, color_low)
  ),
  cmin = min(df$Attendance, na.rm = TRUE),
  cmax = max(df$Attendance, na.rm = TRUE),
  showscale = FALSE,
  colorbar = list(
    title = "Attendance",
    x = 1.02,
    len = 0.75
  ),
  symbol = "circle",
  size = 8
),
    text = ~point_text,
    hoverinfo = "text",
    name = "Attendance-colored outliers",
    legendgroup = "attendance_colored_outliers",
    visible = FALSE,
    showlegend = FALSE
  )


# trend line without outliers: always visible
p <- p %>%
  add_lines(
    data = line_no_out,
    x = ~HoursStudied,
    y = ~ExamScore,
    text = line_text_no_out,
    hoverinfo = "text",
    line = list(color = "black", width = 2),
    name = "Trend line (no outliers)",
    showlegend = FALSE
  )

# trend line with outliers: toggleable
p <- p %>%
  add_lines(
    data = line_all,
    x = ~HoursStudied,
    y = ~ExamScore,
    text = line_text_all,
    hoverinfo = "text",
    line = list(color = "red", width = 2, dash = "dash"),
    name = "Trend line (with outliers)",
    legendgroup = "with_outliers_plain",
    visible = "legendonly",
    showlegend = FALSE
  )

p %>%
  layout(
    title = "Impact of Study Time on Exam Performance and the Role of Attendance",
    xaxis = list(title = "Hours Studied"),
    yaxis = list(title = "Exam Score"),
    dragmode = "zoom",
    legend = list(
      title = list(text = "Click to toggle"),
      groupclick = "togglegroup",
      x = 1.18,
      y = 0.85
    )
  ) %>%
  config(
    displayModeBar = TRUE,
    scrollZoom = TRUE
  )
```

### Key Insights {width=30%}

::: {.card title="Key Insights"}
**What This Shows**

Explores the relationship between hours studied and exam performance, and how attendance modifies this trend.

**Interactive Features**

- Toggle **Color by Attendance** in the legend to color points by attendance rate.
- Toggle **Outliers** to include or exclude extreme data points and see how they shift the trend line.
- **Hover** over the data points and trend line to see the exact scores, slope, correlation, and R² value.
- Select a region to **zoom** in on.

**Study Time & Performance**

- There is a slight positive relationship between hours studied and exam score.
- The shallow slope and low R² suggest that hours studied alone explains only a small portion of score variation.

**The Role of Attendance**

- For the same number of hours studied, students with higher attendance tend to achieve higher exam scores.
- This suggests that **attendance amplifies the benefit of studying**. Time spent in class and time spent studying work together rather than independently.

**Key Takeaway**

Attendance plays an important role in academic success. Showing up consistently appears to matter as much as, or more than, the raw number of hours studied.
:::

## Row {height=500}

### Chart {width=70%}

```{r}
df <- read.csv("StudentPerformanceFactors.csv")
df <- df[df$Distance_from_Home %in% c("Near", "Moderate", "Far"), ]
df <- rbind(
  df[df$Distance_from_Home != "Near", ],
  df[df$Distance_from_Home == "Near", ][sample(1:3884, 1000), ]
)
# Took a sample of the near students because it was overwhelmingly more students on the chart.
p <- ggplot(df, aes(x = Sleep_Hours, y = Exam_Score)) +
  geom_jitter(aes(color = Distance_from_Home), alpha = 0.4, width = 0.25, height = 0.5, size = 1) +
  geom_smooth(aes(color = "Regression Line"), method = "lm", se = TRUE, linewidth = 1.2) +
  scale_color_manual(values = c("Near" = color_high, "Moderate" = color_med, "Far" = color_low, "Regression Line" = color_base), name = "Distance from School") +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = color_base, show.legend = FALSE) +
  scale_y_continuous(limits = c(55, 85), breaks = seq(55, 85, 5)) +
  scale_x_continuous(breaks = 4:10) +
  labs(
    title = "Sleep Hours vs. Exam Score",
    subtitle = "Red line = line of best fit (linear regression). Diamonds = group averages.",
    caption = "The nearly flat regression line indicates little to no relationship between sleep hours and exam scores.",
    x = "Sleep Hours per Night",
    y = "Exam Score (0-100)"
  ) +
  theme_minimal(base_size = 13)
pl <- ggplotly(p)
for (i in seq_along(pl$x$data)) {
  nm <- pl$x$data[[i]]$name
  if (!is.null(nm) && grepl("Regression Line", nm)) {
    pl$x$data[[i]]$visible <- "legendonly"
  }
}
# AI assistance (Claude) was used to brainstorm interactivity ideas for the visualization and to find information on plotly syntax. All design decisions and analysis are my own. Cited as per syllabus.
pl <- pl %>% layout(updatemenus = list(
  list(
    type = "dropdown", x = 1.05, y = 0.5, xanchor = "left",
    buttons = list(
      list(method = "relayout", label = "Zoom: 55-85 (default)",
           args = list(list(yaxis.range = c(55, 85)))),
      list(method = "relayout", label = "Zoom: 60-80 (tight)",
           args = list(list(yaxis.range = c(60, 80)))),
      list(method = "relayout", label = "Full range: 40-100",
           args = list(list(yaxis.range = c(40, 100))))
    )
  ),
  list(
    type = "dropdown", x = 1.05, y = 0.4, xanchor = "left",
    buttons = list(
      list(method = "relayout", label = "Sleep: All (4-10)",
           args = list(list(xaxis.range = c(3.5, 10.5)))),
      list(method = "relayout", label = "Sleep: Low (4-6)",
           args = list(list(xaxis.range = c(3.5, 6.5)))),
      list(method = "relayout", label = "Sleep: Normal (7-8)",
           args = list(list(xaxis.range = c(6.5, 8.5)))),
      list(method = "relayout", label = "Sleep: High (9-10)",
           args = list(list(xaxis.range = c(8.5, 10.5))))
    )
  )
  
),
legend = list(title = list(text = "<b>Click to filter</b><br>Distance from School"))
)
pl
```

### Key Insights {width=30%}

::: {.card title="Key Insights"}
**What This Shows**

Explores the relationship between sleep hours and exam scores, with distance from school as a secondary dimension to test whether location changes the pattern.

**Interactive Features**

- Toggle **Regression Line** in the legend to overlay the line of best fit.
- Toggle individual distance groups (Near, Moderate, Far) on and off to isolate specific subsets.
- Hover over points to see individual student values.
- The red diamonds mark the average exam score at each sleep hour level.
- Dropdowns to shrink Axis Scale and to pick different subsets of sleep hours.

**Sleep & Performance**

- There is little to no correlation between sleep hours and exam scores and the regression line is nearly flat across the full 4–10 hour range.
- Group averages (red diamonds) remain stable across sleep levels, reinforcing the lack  of a meaningful trend.

**The Role of Distance**

- Students near, moderate, and far from school show similar score distributions at every sleep level.
- Distance from school does not appear to modify or strengthen the sleep-score relationship.

**Key Takeaway**

Sleep alone is not a strong driver of academic performance in this dataset. The flat trend holds across all distance groups, suggesting other factors carry more explanatory weight.
:::

## Row {height=600}

### Chart {width=70%}

```{r}
students <- read.csv("StudentPerformanceFactors.csv")

profile_stats <- students %>%
  mutate(
    is_athlete = ifelse(Physical_Activity >= 4, "Athlete", "Non-Athlete"),
    has_extra = ifelse(Extracurricular_Activities == "Yes", "+ Extras", "No Extras")
  ) %>%
  group_by(is_athlete, has_extra) %>%
  summarize(
    study_hrs = round(mean(Hours_Studied, na.rm = TRUE), 1),
    sleep_hrs = round(mean(Sleep_Hours, na.rm = TRUE), 1),
    exam_score = round(mean(Exam_Score, na.rm = TRUE), 1),
    .groups = 'drop'
  ) %>%
  mutate(
    # UPDATED: Shortened the prefix to Ath. and Non-Ath. for the bar labels
    profile_name = paste0(ifelse(is_athlete == "Athlete", "Ath.", "Non-Ath."), "<br>", has_extra)
  )

gap_row <- data.frame(
  is_athlete = "Gap",
  has_extra = "Gap",
  profile_name = " ", 
  study_hrs = NA,
  sleep_hrs = NA,
  exam_score = NA
)

plot_data <- bind_rows(
  profile_stats %>% filter(is_athlete == "Athlete"),
  gap_row,
  profile_stats %>% filter(is_athlete == "Non-Athlete")
)

ath <- profile_stats %>% filter(is_athlete == "Athlete")
non <- profile_stats %>% filter(is_athlete == "Non-Athlete")

sort_study_desc <- c(ath %>% arrange(desc(study_hrs)) %>% pull(profile_name), " ", non %>% arrange(desc(study_hrs)) %>% pull(profile_name))
sort_study_asc  <- c(ath %>% arrange(study_hrs) %>% pull(profile_name), " ", non %>% arrange(study_hrs) %>% pull(profile_name))
sort_sleep_desc <- c(ath %>% arrange(desc(sleep_hrs)) %>% pull(profile_name), " ", non %>% arrange(desc(sleep_hrs)) %>% pull(profile_name))
sort_sleep_asc  <- c(ath %>% arrange(sleep_hrs) %>% pull(profile_name), " ", non %>% arrange(sleep_hrs) %>% pull(profile_name))
sort_score_desc <- c(ath %>% arrange(desc(exam_score)) %>% pull(profile_name), " ", non %>% arrange(desc(exam_score)) %>% pull(profile_name))
sort_score_asc  <- c(ath %>% arrange(exam_score) %>% pull(profile_name), " ", non %>% arrange(exam_score) %>% pull(profile_name))

plot_data$profile_name <- factor(plot_data$profile_name, levels = sort_study_desc)

time_plot <- plot_ly(plot_data, x = ~profile_name) %>%
  add_bars(
    y = ~study_hrs,
    name = "Study Hours",
    marker = list(color = color_high), 
    hovertemplate = "<b>%{x}</b><br>Study: %{y} hrs<extra></extra>"
  ) %>%
  add_bars(
    y = ~sleep_hrs,
    name = "Sleep Hours",
    marker = list(color = color_low), 
    hovertemplate = "<b>%{x}</b><br>Sleep: %{y} hrs<extra></extra>"
  ) %>%
  layout(
    yaxis = list(title = "Average Hours", gridcolor = "#e5e5e5", range = c(0, 22)),
    barmode = 'group'
  )

score_plot <- plot_ly(plot_data, x = ~profile_name) %>%
  add_bars(
    y = ~exam_score,
    name = "Final Exam Score",
    marker = list(color = color_med), 
    hovertemplate = "<b>%{x}</b><br>Exam Score: %{y}<extra></extra>"
  ) %>%
  layout(
    yaxis = list(title = "Average Score", gridcolor = "#e5e5e5", range = c(0, 110))
  )

sorting_dropdown <- list(
  type = "dropdown",
  direction = "down",
  x = 1,
  xref = "container",
  xanchor = "right",
  y = 0.98,
  yref = "container",
  yanchor = "top",
  buttons = list(
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_study_desc, "xaxis2.categoryorder" = "array", "xaxis2.categoryarray" = sort_study_desc)), label = "Sort: Study (High to Low)"),
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_study_asc, "xaxis2.categoryorder" = "array", "xaxis2.categoryarray" = sort_study_asc)), label = "Sort: Study (Low to High)"),
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_sleep_desc, "xaxis2.categoryorder" = "array", "xaxis2.categoryarray" = sort_sleep_desc)), label = "Sort: Sleep (High to Low)"),
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_sleep_asc, "xaxis2.categoryorder" = "array", "xaxis2.categoryarray" = sort_sleep_asc)), label = "Sort: Sleep (Low to High)"),
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_score_desc, "xaxis2.categoryorder" = "array", "xaxis2.categoryarray" = sort_score_desc)), label = "Sort: Score (High to Low)"),
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_score_asc, "xaxis2.categoryorder" = "array", "xaxis2.categoryarray" = sort_score_asc)), label = "Sort: Score (Low to High)")
  )
)

final_plot <- subplot(time_plot, score_plot, nrows = 1, margin = 0.05, titleX = FALSE, titleY = TRUE) %>%
  layout(
    title = list(
      text = "<b>Impact of Athletics & Extracurriculars<br>on Habits and Performance</b>",
      font = list(size = 16, color = "#232D4B"),
      x = 0.5,
      xanchor = "center",
      y = 0.93,
      yanchor = "top",
      yref = "container"
    ),
    annotations = list(
      # Original top headers
      list(x = 0.24, y = 1.02, text = "<b>Study and Sleep Time</b>", showarrow = FALSE, xref = "paper", yref = "paper", font = list(size = 14), xanchor = "center", yanchor = "bottom"),
      list(x = 0.76, y = 1.02, text = "<b>Exam Scores</b>", showarrow = FALSE, xref = "paper", yref = "paper", font = list(size = 14), xanchor = "center", yanchor = "bottom"),
      
      # ALIGNMENT/PADDING FIX: Tweak the "x" decimals below to nudge the labels left or right!
      # Decrease "y" (e.g., -0.3) to push them further down.
      list(x = 0.05, y = -0.25, text = "<b>ATHLETE</b>", showarrow = FALSE, xref = "paper", yref = "paper", font = list(size = 12, color = color_base)),
      list(x = 0.36, y = -0.25, text = "<b>NON-ATHLETE</b>", showarrow = FALSE, xref = "paper", yref = "paper", font = list(size = 12, color = color_base)),
      list(x = 0.63, y = -0.25, text = "<b>ATHLETE</b>", showarrow = FALSE, xref = "paper", yref = "paper", font = list(size = 12, color = color_base)),
      list(x = 0.98, y = -0.25, text = "<b>NON-ATHLETE</b>", showarrow = FALSE, xref = "paper", yref = "paper", font = list(size = 12, color = color_base))
    ),
    legend = list(orientation = "h", x = 0.5, xanchor = "center", y = -0.4),
    margin = list(t = 130, b = 150), 
    paper_bgcolor = "white",
    plot_bgcolor = "white",
    barmode = 'group',
    xaxis = list(tickangle = -45, tickfont = list(size = 11)),
    xaxis2 = list(tickangle = -45, tickfont = list(size = 11)),
    updatemenus = list(sorting_dropdown)
  )

final_plot
```

### Key Insights {width=30%}

::: {.card title="Key Insights"}
**What This Shows**

Compares average study hours, sleep hours, and exam scores across four student profiles defined by athletic involvement and extracurricular participation.

**Interactive Features**

- Use the **sort dropdown** to reorder profiles by study hours, sleep hours, or exam score.
- Hover over any bar to see the exact average value for that profile.
- Toggle **Study Hours** or **Sleep Hours** in the legend to isolate individual metrics.

**Habits Across Profiles**

- Differences in study and sleep hours across the four profiles are small, suggesting that athletic and extracurricular involvement does not hurt daily routines much.
- Athletes with extracurriculars study slightly fewer hours on average, yet their exam scores remain comparable to less-involved peers.

**Performance Outcomes**

- Exam scores are nearly identical across all four profiles, regardless of involvement level.
- Non-athletes with no extracurriculars study the most hours but show no corresponding score advantage, hinting at diminishing returns from isolated study.

**Key Takeaway**

Participation in sports and extracurriculars does not correlate with lower academic performance. Highly involved students appear to manage their time effectively, balancing external commitments without sacrificing grades. Involvement should be encouraged so students can pursue diverse interests without fearing academic drawback.
:::