---
title: "Student Performance Factors Dashboard"
format: 
  dashboard:
    scrolling: true
    theme: flatly
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
css: styles.css
---

```{r setup, include=FALSE}
library(tidyverse)
library(plotly)
library(scales)
library(DT)
library(ggplot2)

#Load data
df <- read.csv("StudentPerformanceFactors.csv")
```

# Overview 
## Row {height= 220}
::: {.card title="Project Overview"}
This dashboard explores the *Student Performance Factors* dataset from Kaggle, which contains information about students' academic performance and the personal and social-economical factors that may influence their exam score.
The dataset captures multiple dimensions of student experience. These include study behaviors (such as hours studied and attendance), lifestyle patterns (including sleep and physical activity), family background (such as income and parental involvement), and school-related conditions (like teacher quality and access to resources). Together, these variables provide a more holistic view of the factors that may influence academic outcomes, with **Exam_Score** as the primary measure of performance.
:::

::: {.card title="Research Questions"}
More specifically, this dashboard aims to:

- Investigate the impact of **support systems**, including parental involvement and access to resources  

- Examine whether **academic habits** (such as hours studied and attendance) are strong predictors of performance  

- Explore how **lifestyle habits** (such as sleep duration) differentiate the hihest and lowest-performing students

- Assess how **extracurricular involvement** (like athletics and clubs) impacts a student's allocation of sleep and study time
:::


## Row
### Static insight: Distribution of Exam Scores {width=50%}
```{r}
ggplot(df, aes(x = Exam_Score)) +
  geom_histogram(bins = 25, fill = "#5C7F92", color = "white") +
  labs(
    title = "Distribution of Exam Scores",
    subtitle = "Scores are roughly normally distributed, centered around 65-70.",
    x = "Exam Score",
    y = "Number of Students"
  ) +
  theme_minimal(base_size = 13)
```


### Static insight {width=50%}
```{r}
motivation_summary <- df %>%
  group_by(Motivation_Level) %>%
  summarise(avg_score = mean(Exam_Score, na.rm = TRUE), .groups = "drop")

motivation_summary$Motivation_Level <- factor(
  motivation_summary$Motivation_Level,
  levels = c("Low", "Medium", "High")
)

ggplot(motivation_summary, aes(x = Motivation_Level, y = avg_score, fill = Motivation_Level)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = round(avg_score, 1)), vjust = -0.4) +
  scale_fill_manual(values = c(
    "Low" = "red",
    "Medium" = "blue",
    "High" = "green"
  )) +
  labs(
    title = "Average Exam Score by Motivation Level",
    x = "Motivation Level",
    y = "Average Exam Score"
  ) +
  theme_minimal(base_size = 13)
```


## Row {height=220}
::: {.card title="What the Distribution Shows"}
The distribution of exam scores provides a baseline view of student performance. Most students cluster within a central range, rather than at the extremes, suggesting that differences in performance are systematic rather than random. This motivates further exploration into which factors may explain this variation.
:::


::: {.card title="What the Motivation Chart Shows"}
Average exam scores do not vary dramatically across motivation levels. While students with higher motivation show slightly better performance, the differences are relatively small. This suggests that motivation alone may not be a strong explanatory factor, and that other variables (such as study habits or external support) may play a more significant role.
:::

## Row {height=190}
### Goals

::: {.card title="Project Goals"}
By combining multiple perspectives, the dashboard seeks to move beyond simple explanations of success and highlight the complexity of student achievement.
The goal of this project is to explore how different aspects of a student's environment and behavior are associated with exam performance, and to identify which factors appear most influential.
:::

<div>
  <a href="#interactive-analysis" class="floating-next">Go to Interactive Analysis</a>
</div>

# Interactive Analysis

## Row {height=500}

### Chart {width=75%}

```{r}
df$Family_Income        <- as.character(df$Family_Income)
df$Parental_Involvement <- as.character(df$Parental_Involvement)
df$Access_to_Resources  <- as.character(df$Access_to_Resources)

pi_levels  <- c("High", "Medium", "Low")
atr_levels <- c("High", "Medium", "Low")
pi_opts    <- c("All", pi_levels)
atr_opts   <- c("All", atr_levels)

income_colors <- c(
  "Low"    = "#EF553B",
  "Medium" = "#00CC96",
  "High"   = "#636EFA"
)

combos <- expand.grid(pi = pi_opts, atr = atr_opts, stringsAsFactors = FALSE)

fig            <- plot_ly()
visibility_map <- list()
trace_index    <- 0

for (i in seq_len(nrow(combos))) {
  pi_val     <- combos$pi[i]
  atr_val    <- combos$atr[i]
  is_default <- (pi_val == "All" && atr_val == "All")

  sub <- df
  if (pi_val  != "All") sub <- sub[sub$Parental_Involvement == pi_val, ]
  if (atr_val != "All") sub <- sub[sub$Access_to_Resources  == atr_val, ]

  sub$Family_Income <- factor(sub$Family_Income, levels = c("Low", "Medium", "High"))

  for (inc in c("Low", "Medium", "High")) {
    sub_inc <- sub[sub$Family_Income == inc, ]
    scores  <- sub_inc$Exam_Score

    mn  <- round(mean(scores, na.rm = TRUE), 2)
    q1  <- quantile(scores, 0.25, na.rm = TRUE)
    med <- median(scores, na.rm = TRUE)
    q3  <- quantile(scores, 0.75, na.rm = TRUE)
    iqr <- q3 - q1
    lo  <- round(min(scores, na.rm = TRUE), 2)
    hi  <- round(max(scores, na.rm = TRUE), 2)

    hover_text <- paste0(
      "<b>", inc, " Income</b><br>",
      "Max: ", hi,  "<br>",
      "Mean: ", mn, "<br>",
      "Median: ", med, "<br>",
      "Min: ", lo
    )

    fig <- fig %>% add_trace(
      x              = sub_inc$Family_Income,
      y              = scores,
      type           = "box",
      name           = inc,
      marker         = list(color = income_colors[[inc]]),
      line           = list(color = income_colors[[inc]]),
      fillcolor      = paste0(income_colors[[inc]], "55"),
      boxpoints      = "outliers",
      hoverinfo      = "none",
      visible        = is_default,
      legendgroup    = inc,
      showlegend     = (i == 0),
      offsetgroup    = inc,
      alignmentgroup = "fixed"
    )

    trace_index <- trace_index + 1
    visibility_map[[length(visibility_map) + 1]] <- list(
      pi = pi_val, atr = atr_val, trace = trace_index
    )

    fig <- fig %>% add_trace(
      x          = inc,
      y          = med,
      type       = "scatter",
      mode       = "markers",
      name       = inc,
      legendgroup = inc,
      showlegend = FALSE,
      marker     = list(color = "rgba(0,0,0,0)", size = 40),
      hovertemplate = paste0(
        "<b>", inc, " Income</b><br>",
        "Max: ", hi,  "<br>",
        "Mean: ", mn, "<br>",
        "Median: ", med, "<br>",
        "Min: ", lo,
        "<extra></extra>"
      ),
      visible    = is_default
    )

    trace_index <- trace_index + 1
    visibility_map[[length(visibility_map) + 1]] <- list(
      pi = pi_val, atr = atr_val, trace = trace_index
    )
    
    outliers <- scores[scores < q1 - 1.5 * iqr | scores > q3 + 1.5 * iqr]
    if (length(outliers) > 0) {
      fig <- fig %>% add_trace(
        x             = rep(inc, length(outliers)),
        y             = outliers,
        type          = "scatter",
        mode          = "markers",
        name          = inc,
        legendgroup   = inc,
        showlegend    = FALSE,
        marker        = list(color = income_colors[[inc]], size = 6),
        hoverlabel    = list(bgcolor = "#000000", font = list(color = "#ffffff")),
        hovertemplate = paste0(
          "<b>", inc, " Income</b><br>",
          "Outlier: %{y}<br>",
          "<extra></extra>"
        ),
        visible = is_default
      )

      trace_index <- trace_index + 1
      visibility_map[[length(visibility_map) + 1]] <- list(
        pi = pi_val, atr = atr_val, trace = trace_index
      )
    }
  }
}

make_vis <- function(sel_pi, sel_atr) {
  vapply(visibility_map, function(m) m$pi == sel_pi && m$atr == sel_atr, logical(1))
}

pi_buttons <- lapply(pi_opts, function(pv) {
  list(method = "update",
       args   = list(list(visible = as.list(make_vis(pv, "All")))),
       label  = pv)
})

atr_buttons <- lapply(atr_opts, function(av) {
  list(method = "update",
       args   = list(list(visible = as.list(make_vis("All", av)))),
       label  = av)
})

fig <- fig %>%
  layout(
    title = list(
      text    = "Effect of Family Income on Exam Performance",
      x       = 0.5,
      xanchor = "center",
      y       = 0.97,
      yanchor = "top",
      font    = list(size = 16)
    ),
    xaxis = list(
      title         = "Family Income",
      type          = "category",
      categoryorder = "array",
      categoryarray = c("Low", "Medium", "High")
    ),
    yaxis   = list(title = "Exam Score"),
    boxmode = "overlay",
    margin  = list(t = 130, r = 30, b = 60, l = 60),

    updatemenus = list(
      list(
        type        = "dropdown",
        active      = 0,
        x           = 0.01,
        xanchor     = "left",
        y           = 1.10,
        yanchor     = "top",
        buttons     = pi_buttons,
        bgcolor     = "#ffffff",
        bordercolor = "#aaaaaa",
        font        = list(size = 12)
      ),
      list(
        type        = "dropdown",
        active      = 0,
        x           = 0.25,
        xanchor     = "left",
        y           = 1.10,
        yanchor     = "top",
        buttons     = atr_buttons,
        bgcolor     = "#ffffff",
        bordercolor = "#aaaaaa",
        font        = list(size = 12)
      )
    ),

    annotations = list(
      list(
        text      = "<b>Parental Involvement:</b>",
        x = 0,  xref = "paper", xanchor = "left",
        y = 1.17, yref = "paper", yanchor = "top",
        showarrow = FALSE, font = list(size = 12, color = "#333333")
      ),
      list(
        text      = "<b>Access to Resources:</b>",
        x = 0.24, xref = "paper", xanchor = "left",
        y = 1.17, yref = "paper", yanchor = "top",
        showarrow = FALSE, font = list(size = 12, color = "#333333")
      )
    )
  )

fig
```

### Key Insights {width=25%}

::: {.card title="Key Insights"}
**Family Income & Performance**

- Higher family income is associated with slightly higher exam scores, but the distributions overlap considerably across all three groups.

- Use the **Parental Involvement** and **Access to Resources** filters to explore how these support factors interact with income.

- When comparing low income with high parental involvement to high income low parental involvement, a clear advantage can be seen for the low income high involvement showing that while income is an important factor, parental involvement can mitigate this income performance disparity.

**Why This Matters**

- Helps understand how background factors influence success.

- Reveals potential inequality in academic outcomes.
:::

## Row {height=500}

### Chart {width=75%}

```{r}
df <- read.csv("StudentPerformanceFactors.csv")

df <- df %>%
  rename(
    HoursStudied = Hours_Studied,
    ExamScore = Exam_Score,
    Attendance = Attendance
  )

q1_h <- quantile(df$HoursStudied, 0.25, na.rm = TRUE)
q3_h <- quantile(df$HoursStudied, 0.75, na.rm = TRUE)
iqr_h <- q3_h - q1_h

q1_s <- quantile(df$ExamScore, 0.25, na.rm = TRUE)
q3_s <- quantile(df$ExamScore, 0.75, na.rm = TRUE)
iqr_s <- q3_s - q1_s

df <- df %>%
  mutate(
    is_outlier =
      HoursStudied < q1_h - 1.5 * iqr_h |
      HoursStudied > q3_h + 1.5 * iqr_h |
      ExamScore < q1_s - 1.5 * iqr_s |
      ExamScore > q3_s + 1.5 * iqr_s
  )

df_main <- df %>% filter(!is_outlier)
df_out  <- df %>% filter(is_outlier)

fit_no_outliers <- lm(ExamScore ~ HoursStudied, data = df_main)
slope_no_outliers <- coef(fit_no_outliers)[2]
intercept_no_outliers <- coef(fit_no_outliers)[1]
corr_no_outliers <- cor(df_main$HoursStudied, df_main$ExamScore, use = "complete.obs")
r2_no_outliers <- summary(fit_no_outliers)$r.squared

fit_all <- lm(ExamScore ~ HoursStudied, data = df)
slope_all <- coef(fit_all)[2]
intercept_all <- coef(fit_all)[1]
corr_all <- cor(df$HoursStudied, df$ExamScore, use = "complete.obs")
r2_all <- summary(fit_all)$r.squared

df_main <- df_main %>%
  mutate(
    point_text = paste0(
      "Hours Studied: ", HoursStudied,
      "<br>Exam Score: ", ExamScore,
      "<br>Attendance: ", Attendance,
      "<br>Status: Non-outlier"
    )
  )

df_out <- df_out %>%
  mutate(
    point_text = paste0(
      "Hours Studied: ", HoursStudied,
      "<br>Exam Score: ", ExamScore,
      "<br>Attendance: ", Attendance,
      "<br>Status: Outlier"
    )
  )

x_seq <- seq(
  min(df$HoursStudied, na.rm = TRUE),
  max(df$HoursStudied, na.rm = TRUE),
  length.out = 200
)

line_no_out <- data.frame(
  HoursStudied = x_seq,
  ExamScore = predict(fit_no_outliers, newdata = data.frame(HoursStudied = x_seq))
)

line_text_no_out <- paste0(
  "Trend line (no outliers)",
  "<br>Slope: ", round(slope_no_outliers, 3),
  "<br>Intercept: ", round(intercept_no_outliers, 3),
  "<br>Correlation (r): ", round(corr_no_outliers, 3),
  "<br>R²: ", round(r2_no_outliers, 3)
)

line_all <- data.frame(
  HoursStudied = x_seq,
  ExamScore = predict(fit_all, newdata = data.frame(HoursStudied = x_seq))
)

line_text_all <- paste0(
  "Trend line (with outliers)",
  "<br>Slope: ", round(slope_all, 3),
  "<br>Intercept: ", round(intercept_all, 3),
  "<br>Correlation (r): ", round(corr_all, 3),
  "<br>R²: ", round(r2_all, 3)
)

p <- plot_ly()

p <- p %>%
  add_markers(
    data = df_main,
    x = ~HoursStudied,
    y = ~ExamScore,
    marker = list(color = "forestgreen", symbol = "circle", size = 8),
    text = ~point_text,
    hoverinfo = "text",
    name = "Students",
    legendgroup = "plain_points",
    showlegend = FALSE,
    visible = TRUE
  )

p <- p %>%
  add_markers(
    data = df_out,
    x = ~HoursStudied,
    y = ~ExamScore,
    marker = list(color = "forestgreen", symbol = "circle", size = 8),
    text = ~point_text,
    hoverinfo = "text",
    name = "Outliers",
    legendgroup = "with_outliers_plain",
    visible = "legendonly",
    showlegend = TRUE
  )

p <- p %>%
  add_markers(
    data = df_main,
    x = ~HoursStudied,
    y = ~ExamScore,
    marker = list(
      color = ~Attendance,
      colorscale = list(c(0, "yellow"), c(1, "darkgreen")),
      cmin = min(df$Attendance, na.rm = TRUE),
      cmax = max(df$Attendance, na.rm = TRUE),
      showscale = TRUE,
      colorbar = list(title = "Attendance", x = 1.02, len = 0.75),
      symbol = "circle",
      size = 8
    ),
    text = ~point_text,
    hoverinfo = "text",
    name = "Color by Attendance",
    legendgroup = "attendance_colored",
    visible = "legendonly",
    showlegend = TRUE
  )

p <- p %>%
  add_markers(
    data = df_out,
    x = ~HoursStudied,
    y = ~ExamScore,
    marker = list(
      color = ~Attendance,
      colorscale = list(c(0, "yellow"), c(1, "darkgreen")),
      cmin = min(df$Attendance, na.rm = TRUE),
      cmax = max(df$Attendance, na.rm = TRUE),
      showscale = FALSE,
      symbol = "circle",
      size = 8
    ),
    text = ~point_text,
    hoverinfo = "text",
    name = "Attendance-colored outliers",
    legendgroup = "attendance_colored_outliers",
    visible = FALSE,
    showlegend = FALSE
  )

p <- p %>%
  add_lines(
    data = line_no_out,
    x = ~HoursStudied,
    y = ~ExamScore,
    text = line_text_no_out,
    hoverinfo = "text",
    line = list(color = "black", width = 2),
    name = "Trend line (no outliers)",
    showlegend = FALSE
  )

p <- p %>%
  add_lines(
    data = line_all,
    x = ~HoursStudied,
    y = ~ExamScore,
    text = line_text_all,
    hoverinfo = "text",
    line = list(color = "red", width = 2, dash = "dash"),
    name = "Trend line (with outliers)",
    legendgroup = "with_outliers_plain",
    visible = "legendonly",
    showlegend = FALSE
  )

p %>%
  layout(
    title = "Impact of Study Time on Exam Performance and the Role of Attendance",
    xaxis = list(title = "Hours Studied"),
    yaxis = list(title = "Exam Score"),
    dragmode = "zoom",
    legend = list(
      title = list(text = "Click to toggle"),
      groupclick = "togglegroup",
      x = 1.18,
      y = 0.85
    )
  ) %>%
  config(displayModeBar = TRUE, scrollZoom = TRUE)
```

### Key Insights {width=25%}

::: {.card title="Key Insights"}
**Study Time & Exam Scores**

- There is a modest positive relationship between hours studied and exam score — more study time tends to correlate with slightly higher scores.

- The trend line slope is relatively shallow, indicating that study time alone explains only a small portion of score variation (check the R² by hovering the trend line).

- Toggle **Color by Attendance** to reveal that students with higher attendance (darker green) tend to cluster at higher scores regardless of study hours.

- This suggests **attendance may be a stronger predictor** than raw study time — consistent presence in class compounds the benefit of studying.
:::

## Row {height=500}

### Chart {width=75%}

```{r}
#| echo: false
#| message: false
#| warning: false

df <- read.csv("StudentPerformanceFactors.csv")
df <- df[df$Distance_from_Home %in% c("Near", "Moderate", "Far"), ]
df <- rbind(
  df[df$Distance_from_Home != "Near", ],
  df[df$Distance_from_Home == "Near", ][sample(1:3884, 1000), ]
)

p <- ggplot(df, aes(x = Sleep_Hours, y = Exam_Score)) +
  geom_jitter(aes(color = Distance_from_Home), alpha = 0.4, width = 0.25, height = 0.5, size = 1) +
  geom_smooth(aes(color = "Regression Line"), method = "lm", se = TRUE, linewidth = 1.2) +
  scale_color_manual(values = c("Near" = "#4CAF50", "Moderate" = "#FFC107", "Far" = "#F44336", "Regression Line" = "red"), name = "Distance from School") +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "red", show.legend = FALSE) +
  scale_y_continuous(limits = c(55, 85), breaks = seq(55, 85, 5)) +
  scale_x_continuous(breaks = 4:10) +
  labs(
    title = "Sleep Hours vs. Exam Score",
    subtitle = "Red line = line of best fit (linear regression). Diamonds = group averages.",
    caption = "The nearly flat regression line indicates little to no relationship between sleep hours and exam scores.",
    x = "Sleep Hours per Night",
    y = "Exam Score (0–100)"
  ) +
  theme_minimal(base_size = 13)

pl <- ggplotly(p)
for (i in seq_along(pl$x$data)) {
  nm <- pl$x$data[[i]]$name
  if (!is.null(nm) && grepl("Regression Line", nm)) {
    pl$x$data[[i]]$visible <- "legendonly"
  }
}
pl
```

### Key Insights {width=25%}

::: {.card title="Key Insights"}
**Sleep Hours & Exam Scores**

- The regression line is nearly flat, indicating that sleep hours alone have little to no linear relationship with exam performance in this dataset.

- Score distributions are similar across the 4–10 hour sleep range, suggesting there is no clearly optimal sleep amount for academic performance here.

- Distance from home does not appear to create meaningful clustering — students near, moderate, and far from school show similar score spreads at each sleep level.

- The red diamonds (group averages) remain relatively stable across sleep categories, reinforcing the absence of a strong trend.
:::

## Row {height=600}

### Chart {width=75%}

```{r}
students <- read.csv("StudentPerformanceFactors.csv")

profile_stats <- students %>%
  mutate(
    is_athlete = ifelse(Physical_Activity >= 4, "Athlete", "Non-Athlete"),
    has_extra = ifelse(Extracurricular_Activities == "Yes", "+ Extras", "No Extras")
  ) %>%
  group_by(is_athlete, has_extra) %>%
  summarize(
    study_hrs = round(mean(Hours_Studied, na.rm = TRUE), 1),
    sleep_hrs = round(mean(Sleep_Hours, na.rm = TRUE), 1),
    exam_score = round(mean(Exam_Score, na.rm = TRUE), 1),
    .groups = 'drop'
  ) %>%
  mutate(
    profile_name = paste0(is_athlete, "<br>", has_extra)
  )

gap_row <- data.frame(
  is_athlete = "Gap",
  has_extra = "Gap",
  profile_name = " ", 
  study_hrs = NA,
  sleep_hrs = NA,
  exam_score = NA
)

plot_data <- bind_rows(
  profile_stats %>% filter(is_athlete == "Athlete"),
  gap_row,
  profile_stats %>% filter(is_athlete == "Non-Athlete")
)

ath <- profile_stats %>% filter(is_athlete == "Athlete")
non <- profile_stats %>% filter(is_athlete == "Non-Athlete")

sort_study_desc <- c(ath %>% arrange(desc(study_hrs)) %>% pull(profile_name), " ", non %>% arrange(desc(study_hrs)) %>% pull(profile_name))
sort_study_asc  <- c(ath %>% arrange(study_hrs) %>% pull(profile_name), " ", non %>% arrange(study_hrs) %>% pull(profile_name))
sort_sleep_desc <- c(ath %>% arrange(desc(sleep_hrs)) %>% pull(profile_name), " ", non %>% arrange(desc(sleep_hrs)) %>% pull(profile_name))
sort_sleep_asc  <- c(ath %>% arrange(sleep_hrs) %>% pull(profile_name), " ", non %>% arrange(sleep_hrs) %>% pull(profile_name))
sort_score_desc <- c(ath %>% arrange(desc(exam_score)) %>% pull(profile_name), " ", non %>% arrange(desc(exam_score)) %>% pull(profile_name))
sort_score_asc  <- c(ath %>% arrange(exam_score) %>% pull(profile_name), " ", non %>% arrange(exam_score) %>% pull(profile_name))

plot_data$profile_name <- factor(plot_data$profile_name, levels = sort_study_desc)

time_plot <- plot_ly(plot_data, x = ~profile_name) %>%
  add_bars(
    y = ~study_hrs,
    name = "Study Hours",
    marker = list(color = "#232D4B"),
    hovertemplate = "<b>%{x}</b><br>Study: %{y} hrs<extra></extra>"
  ) %>%
  add_bars(
    y = ~sleep_hrs,
    name = "Sleep Hours",
    marker = list(color = "#E57200"),
    hovertemplate = "<b>%{x}</b><br>Sleep: %{y} hrs<extra></extra>"
  ) %>%
  layout(
    yaxis = list(title = "Average Hours", gridcolor = "#e5e5e5", range = c(0, 22)),
    barmode = 'group'
  )

score_plot <- plot_ly(plot_data, x = ~profile_name) %>%
  add_bars(
    y = ~exam_score,
    name = "Final Exam Score",
    marker = list(color = "#5C7F92"),
    hovertemplate = "<b>%{x}</b><br>Exam Score: %{y}<extra></extra>"
  ) %>%
  layout(
    yaxis = list(title = "Average Score", gridcolor = "#e5e5e5", range = c(0, 110))
  )

sorting_dropdown <- list(
  type = "dropdown",
  direction = "down",
  x = 1,
  xref = "container",
  xanchor = "right",
  y = 0.98,
  yref = "container",
  yanchor = "top",
  buttons = list(
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_study_desc, "xaxis2.categoryorder" = "array", "xaxis2.categoryarray" = sort_study_desc)), label = "Sort: Study (High to Low)"),
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_study_asc, "xaxis2.categoryorder" = "array", "xaxis2.categoryarray" = sort_study_asc)), label = "Sort: Study (Low to High)"),
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_sleep_desc, "xaxis2.categoryorder" = "array", "xaxis2.categoryarray" = sort_sleep_desc)), label = "Sort: Sleep (High to Low)"),
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_sleep_asc, "xaxis2.categoryorder" = "array", "xaxis2.categoryarray" = sort_sleep_asc)), label = "Sort: Sleep (Low to High)"),
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_score_desc, "xaxis2.categoryorder" = "array", "xaxis2.categoryarray" = sort_score_desc)), label = "Sort: Score (High to Low)"),
    list(method = "relayout", args = list(list("xaxis.categoryorder" = "array", "xaxis.categoryarray" = sort_score_asc, "xaxis2.categoryorder" = "array", "xaxis2.categoryarray" = sort_score_asc)), label = "Sort: Score (Low to High)")
  )
)

final_plot <- subplot(time_plot, score_plot, nrows = 1, margin = 0.05, titleX = FALSE, titleY = TRUE) %>%
  layout(
    title = list(
      text = "<b>Impact of Athletics & Extracurriculars<br>on Habits and Performance</b>",
      font = list(size = 16, color = "#232D4B"),
      x = 0.5,
      xanchor = "center",
      y = 0.93,
      yanchor = "top",
      yref = "container"
    ),
    annotations = list(
      list(x = 0.24, y = 1.02, text = "<b>Study and Sleep Time</b>", showarrow = FALSE, xref = "paper", yref = "paper", font = list(size = 14), xanchor = "center", yanchor = "bottom"),
      list(x = 0.76, y = 1.02, text = "<b>Exam Scores</b>", showarrow = FALSE, xref = "paper", yref = "paper", font = list(size = 14), xanchor = "center", yanchor = "bottom")
    ),
    legend = list(orientation = "h", x = 0.5, xanchor = "center", y = -0.35),
    margin = list(t = 130, b = 120),
    paper_bgcolor = "white",
    plot_bgcolor = "white",
    barmode = 'group',
    xaxis = list(tickangle = -45, tickfont = list(size = 11)),
    xaxis2 = list(tickangle = -45, tickfont = list(size = 11)),
    updatemenus = list(sorting_dropdown)
  )

final_plot
```

### Key Insights {width=25%}

::: {.card title="Key Insights"}
**Athletics, Extracurriculars & Performance**

- Across all four profiles, differences in average exam scores are small, suggesting that being an athlete or joining extracurriculars does not strongly hurt or help performance.

- Athletes with extracurriculars study slightly fewer hours on average, yet their exam scores remain comparable — hinting at efficiency or time-management advantages.

- Non-athletes with no extracurriculars study the most hours but do not show correspondingly higher scores, which may suggest diminishing returns on isolated study.

- Use the **sort dropdown** to reorder profiles by study, sleep, or score and see how the tradeoffs shift across groups.
:::