---
title: "Visualization 2"
format: 
  dashboard:
    orientation: rows
    theme: cosmo
    embed-resources: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(plotly)

library(ggplot2)
library(scales)
library(dplyr)
```



## Row 
### Visualization 2

```{r}
df <- read.csv("StudentPerformanceFactors.csv")

# rename columns
df <- df %>%
  rename(
    HoursStudied = Hours_Studied,
    ExamScore = Exam_Score,
    Attendance = Attendance
  )

# filter outliers
q1_h <- quantile(df$HoursStudied, 0.25, na.rm = TRUE)
q3_h <- quantile(df$HoursStudied, 0.75, na.rm = TRUE)
iqr_h <- q3_h - q1_h

q1_s <- quantile(df$ExamScore, 0.25, na.rm = TRUE)
q3_s <- quantile(df$ExamScore, 0.75, na.rm = TRUE)
iqr_s <- q3_s - q1_s

df <- df %>%
  mutate(
    is_outlier =
      HoursStudied < q1_h - 1.5 * iqr_h |
      HoursStudied > q3_h + 1.5 * iqr_h |
      ExamScore < q1_s - 1.5 * iqr_s |
      ExamScore > q3_s + 1.5 * iqr_s
  )

df_main <- df %>% filter(!is_outlier)
df_out  <- df %>% filter(is_outlier)

# fit regression line (without outliers)
fit_no_outliers <- lm(ExamScore ~ HoursStudied, data = df_main)

slope_no_outliers <- coef(fit_no_outliers)[2]
intercept_no_outliers <- coef(fit_no_outliers)[1]
corr_no_outliers <- cor(df_main$HoursStudied, df_main$ExamScore, use = "complete.obs")
r2_no_outliers <- summary(fit_no_outliers)$r.squared

# regression with all data
fit_all <- lm(ExamScore ~ HoursStudied, data = df)

slope_all <- coef(fit_all)[2]
intercept_all <- coef(fit_all)[1]
corr_all <- cor(df$HoursStudied, df$ExamScore, use = "complete.obs")
r2_all <- summary(fit_all)$r.squared

# hover text
df_main <- df_main %>%
  mutate(
    point_text = paste0(
      "Hours Studied: ", HoursStudied,
      "<br>Exam Score: ", ExamScore,
      "<br>Attendance: ", Attendance,
      "<br>Status: Non-outlier"
    )
  )

df_out <- df_out %>%
  mutate(
    point_text = paste0(
      "Hours Studied: ", HoursStudied,
      "<br>Exam Score: ", ExamScore,
      "<br>Attendance: ", Attendance,
      "<br>Status: Outlier"
    )
  )

# predicted x values for regression lines
x_seq <- seq(
  min(df$HoursStudied, na.rm = TRUE),
  max(df$HoursStudied, na.rm = TRUE),
  length.out = 200
)

# line without outliers
line_no_out <- data.frame(
  HoursStudied = x_seq,
  ExamScore = predict(fit_no_outliers, newdata = data.frame(HoursStudied = x_seq))
)

line_text_no_out <- paste0(
  "Trend line (no outliers)",
  "<br>Slope: ", round(slope_no_outliers, 3),
  "<br>Intercept: ", round(intercept_no_outliers, 3),
  "<br>Correlation (r): ", round(corr_no_outliers, 3),
  "<br>R²: ", round(r2_no_outliers, 3)
)

# line with all data
line_all <- data.frame(
  HoursStudied = x_seq,
  ExamScore = predict(fit_all, newdata = data.frame(HoursStudied = x_seq))
)

line_text_all <- paste0(
  "Trend line (with outliers)",
  "<br>Slope: ", round(slope_all, 3),
  "<br>Intercept: ", round(intercept_all, 3),
  "<br>Correlation (r): ", round(corr_all, 3),
  "<br>R²: ", round(r2_all, 3)
)

p <- plot_ly()

# -----------------------------
# DEFAULT VIEW: plain green
# -----------------------------

# main data, green, visible by default
p <- p %>%
  add_markers(
    data = df_main,
    x = ~HoursStudied,
    y = ~ExamScore,
    marker = list(
      color = "forestgreen",
      symbol = "circle",
      size = 8
    ),
    text = ~point_text,
    hoverinfo = "text",
    name = "Students",
    legendgroup = "plain_points",
    showlegend = FALSE,
    visible = TRUE
  )

# outliers, green, hidden by default
p <- p %>%
  add_markers(
    data = df_out,
    x = ~HoursStudied,
    y = ~ExamScore,
    marker = list(
      color = "forestgreen",
      symbol = "circle",
      size = 8
    ),
    text = ~point_text,
    hoverinfo = "text",
    name = "Outliers",
    legendgroup = "with_outliers_plain",
    visible = "legendonly",
    showlegend = TRUE
  )

# -----------------------------
# ATTENDANCE-COLORED VIEW
# -----------------------------

# main data colored by attendance, hidden by default
p <- p %>%
  add_markers(
    data = df_main,
    x = ~HoursStudied,
    y = ~ExamScore,
    marker = list(
  color = ~Attendance,
  colorscale = list(
    c(0, "yellow"),
    c(1, "darkgreen")
  ),
  cmin = min(df$Attendance, na.rm = TRUE),
  cmax = max(df$Attendance, na.rm = TRUE),
  showscale = TRUE,
  colorbar = list(
    title = "Attendance",
    x = 1.02,
    len = 0.75
  ),
  symbol = "circle",
  size = 8
),
    text = ~point_text,
    hoverinfo = "text",
    name = "Color by Attendance",
    legendgroup = "attendance_colored",
    visible = "legendonly",
    showlegend = TRUE
  )

# outliers colored by attendance, hidden by default
p <- p %>%
  add_markers(
    data = df_out,
    x = ~HoursStudied,
    y = ~ExamScore,
    marker = list(
  color = ~Attendance,
  colorscale = list(
    c(0, "yellow"),
    c(1, "darkgreen")
  ),
  cmin = min(df$Attendance, na.rm = TRUE),
  cmax = max(df$Attendance, na.rm = TRUE),
  showscale = FALSE,
  colorbar = list(
    title = "Attendance",
    x = 1.02,
    len = 0.75
  ),
  symbol = "circle",
  size = 8
),
    text = ~point_text,
    hoverinfo = "text",
    name = "Attendance-colored outliers",
    legendgroup = "attendance_colored_outliers",
    visible = FALSE,
    showlegend = FALSE
  )

# -----------------------------
# TREND LINES
# -----------------------------

# trend line without outliers: always visible
p <- p %>%
  add_lines(
    data = line_no_out,
    x = ~HoursStudied,
    y = ~ExamScore,
    text = line_text_no_out,
    hoverinfo = "text",
    line = list(color = "black", width = 2),
    name = "Trend line (no outliers)",
    showlegend = FALSE
  )

# trend line with outliers: toggleable
p <- p %>%
  add_lines(
    data = line_all,
    x = ~HoursStudied,
    y = ~ExamScore,
    text = line_text_all,
    hoverinfo = "text",
    line = list(color = "red", width = 2, dash = "dash"),
    name = "Trend line (with outliers)",
    legendgroup = "with_outliers_plain",
    visible = "legendonly",
    showlegend = FALSE
  )

# -----------------------------
# BUTTONS
# -----------------------------
# Trace order:
# 1 plain main
# 2 plain outliers
# 3 attendance main
# 4 attendance outliers
# 5 no-outlier trend
# 6 all-data trend

p %>%
  layout(
    title = "Impact of Study Time on Exam Performance and the Role of Attendance",
    xaxis = list(title = "Hours Studied"),
    yaxis = list(title = "Exam Score"),
    dragmode = "zoom",
    legend = list(
      title = list(text = "Click to toggle"),
      groupclick = "togglegroup",
      x = 1.18,
      y = 0.85
    )
  ) %>%
  config(
    displayModeBar = TRUE,
    scrollZoom = TRUE
  )
```

## Row
### Interpretation
There is a slight positive relationship between hours studied and exam score, suggesting that increased study time is generally associated with better performance. Additionally, for the same number of hours studied, students with higher attendance tend to achieve higher exam scores, indicating that attendance may play an important supporting role in academic success.